---
title: "16S Microbiome Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: united
knit:
  root.dir: "D:/OneDrive - Emory/Research/Jones R/Mouse Data/MS_10_Antipsychotic/16S Analysis/Nephele_analysis/r_analysis_2"
---

Find Files
```{r}
seqtab <- readRDS("seqtab_nochimera.rds")
taxa <- as.matrix(read.table("taxonomy_table.txt", header = TRUE, row.names = 1, sep = "\t"))
tree <- read_tree("rooted_tree.nwk")
tree$tip.label <- colnames(seqtab)
rownames(taxa) <- colnames(seqtab)

metadata <- read.table(
  "mapping_file.txt",
  header = TRUE,
  row.names = 1,
  sep = "\t",
  comment.char = "",
  check.names = FALSE
)

common_samples <- intersect(rownames(seqtab), rownames(metadata))
seqtab <- seqtab[common_samples, ]
metadata <- metadata[common_samples, ]

metadata <- metadata %>%
  rownames_to_column("SampleID") %>%
  mutate(across(everything(), ~ trimws(as.character(.)))) %>%
  column_to_rownames("SampleID")

ps <- phyloseq(
  otu_table(seqtab, taxa_are_rows = FALSE),
  tax_table(taxa),
  sample_data(metadata),
  phy_tree(tree)
)

# Validation checks
stopifnot(identical(sort(rownames(seqtab)), sort(rownames(metadata))))
stopifnot(identical(sort(colnames(seqtab)), sort(rownames(taxa))))
stopifnot(identical(sort(tree$tip.label), sort(colnames(seqtab))))

sample_sums(ps) %>% summary()

```

---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```
---

### 🧩 Conditional Package Installation Chunk


```{r install-packages}
cran_packages <- c(
  "ggplot2", "vegan", "patchwork", "data.table", "dplyr", "RColorBrewer",
  "tidyr", "readr", "gridExtra", "knitr", "stringr", "ggpubr", "ggthemes", "tibble"
)
bioc_packages <- c("phyloseq", "DESeq2", "ape")

install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}
invisible(lapply(cran_packages, install_if_missing))

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
invisible(lapply(bioc_packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg, ask = FALSE, update = FALSE)
}))


```

```{r load-libraries}
library(phyloseq)
library(ggplot2)
library(vegan)
library(DESeq2)
library(ape)
library(patchwork)
library(data.table)
library(dplyr)
library(RColorBrewer)
library(tidyr)
library(readr)
library(knitr)
library(stringr)
library(ggpubr)
library(ggthemes)
library(tibble)
library(broom)
library(phyloseq)
library(ggplot2)
library(patchwork)
library(ggforce)
library(vegan)
library(mixOmics)
library(tibble)
library(dplyr)

```

Set Working Directories 
```{r setup, include=FALSE}
library(knitr)
# Set this to where your .rds files are
root.dir <- "D:/OneDrive - Emory/Research/Jones R/Mouse Data/MS_10_Antipsychotic/16S Analysis/Nephele_analysis/r_analysis_2"
knitr::opts_knit$set(root.dir = root.dir)
setwd(knitr::opts_knit$get("root.dir")[[1]])


```


R.utils::gunzip("seqtab_nochimera.rds", destname = "seqtab_unzipped.rds", overwrite = TRUE)
```{r load-seqtab}
getwd()
list.files()
```
Set COlor pallette 
```{r}
NaturePastel_palette <- c(
  "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e",
  "#e6ab02", "#a6761d", "#666666", "#1f78b4", "#b2df8a",
  "#fb9a99", "#fdbf6f", "#cab2d6", "#ffff99", "#b15928",
  "#8dd3c7", "#bebada", "#fb8072", "#80b1d3", "#fdb462"
)
```

## 1. Load DAD2 Outputs
```{r}

seqtab <- readRDS("seqtab_unzipped.rds")

taxa <- as.matrix(read.table("taxonomy_table.txt", header = TRUE, row.names = 1, sep = "\t"))
tree <- read_tree("rooted_tree.nwk")
tree$tip.label <- colnames(seqtab)
rownames(taxa) <- colnames(seqtab)

metadata <- read.table(
  "mapping_file.txt",
  header = TRUE,
  row.names = 1,
  sep = "\t",
  comment.char = "",
  check.names = FALSE
)

common_samples <- intersect(rownames(seqtab), rownames(metadata))
seqtab <- seqtab[common_samples, ]
metadata <- metadata[common_samples, ]

metadata <- metadata %>%
  rownames_to_column("SampleID") %>%
  mutate(across(everything(), ~ trimws(as.character(.)))) %>%
  column_to_rownames("SampleID")

ps <- phyloseq(
  otu_table(seqtab, taxa_are_rows = FALSE),
  tax_table(taxa),
  sample_data(metadata),
  phy_tree(tree)
)

# Validation checks
stopifnot(identical(sort(rownames(seqtab)), sort(rownames(metadata))))
stopifnot(identical(sort(colnames(seqtab)), sort(rownames(taxa))))
stopifnot(identical(sort(tree$tip.label), sort(colnames(seqtab))))

sample_sums(ps) %>% summary()

```

Save Phyloseq object
```{r}
saveRDS(ps, file = "phyloseq_object.rds")

```

```{r alpha_diversity_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8}

# Prepare richness data including Simpson
richness_raw <- estimate_richness(ps, measures = c("Shannon", "Observed", "Simpson")) %>%
  rownames_to_column("SampleID")

# Load metadata
metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")

# Merge and reshape
richness_df <- full_join(metadata_df, richness_raw, by = "SampleID") %>%
  pivot_longer(cols = c("Shannon", "Observed", "Simpson"), names_to = "measure", values_to = "value")

# Filter and re-level time points
richness_df <- richness_df %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6"))

richness_df$Time_Point <- factor(richness_df$Time_Point, levels = c("T0", "T2", "T3", "T6"))

# Define consistent TreatmentGroup order
richness_df$TreatmentGroup <- factor(richness_df$TreatmentGroup, 
                                     levels = unique(metadata_df$TreatmentGroup))

# Updated consistent color palette
group_colors <- c(
  "Vehicle"     = "#FFFFFF",
  "Haloperidol" = "#62666A",
  "Quetiapine"  = "#E9E4DC",
  "Olanzapine"  = "#81BDD1",
  "Risperidone" = "#C3AC84"
)

# Relevel TreatmentGroup according to color order
richness_df$TreatmentGroup <- factor(
  richness_df$TreatmentGroup,
  levels = names(group_colors)
)


# Generate the plot
p <- ggplot(richness_df, aes(x = Time_Point, y = value, fill = TreatmentGroup)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, position = position_dodge(width = 0.8)) +
  geom_jitter(
    aes(color = TreatmentGroup),
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1,
    alpha = 0.5
  ) +
  geom_smooth(
    aes(group = TreatmentGroup, color = TreatmentGroup),
    method = "lm",
    se = FALSE,
    linetype = "dashed"
  ) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  theme_minimal() +
  labs(
    title = "Alpha Diversity Over Time by Treatment Group",
    x = "Time Point",
    y = "Diversity Index"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# Display the plot
print(p)

# Save the plot
ggsave(filename = file.path(root.dir, "alpha_diversity_plot_2_SampsonColors.tiff"), plot = p, width = 16, height = 8, dpi = 300)
ggsave(filename = file.path(root.dir, "alpha_diversity_plot_2_SampsonColors.pdf"), plot = p, width = 16, height = 8)
write.csv(richness_df,
          file = file.path(root.dir, "richness_df_for_alpha_plot.csv"),
          row.names = FALSE)

```
can we simplify these plots a little to make them easier to interpret? Shannon and Simpson have relatively similar trends. can we show just Observed and either Shannon or Inverse Simpson? Also instead of bar plots, can we tryline plots with mean +/- SE of the mean or 95% CI with solid lines for the means and semi-transparent error bands?
computes Observed, Shannon, and InvSimpson (where InvSimpson = 1/Simpson from estimate_richness)

filters to T0, T2, T3, T6

summarizes per TreatmentGroup × Time_Point × measure

plots solid mean lines + semi-transparent ribbons (choose SE or 95% CI)
ribbon_type <- "SE"

facets only Observed + Shannon or Observed + InvSimpson
```{r alpha_diversity_lineplot_simplified, echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

# ---- settings ----
keep_time <- c("T0", "T2", "T3", "T6")

# Choose ONE of: "Shannon" or "InvSimpson"
evenness_metric <- "Shannon"   # change to "InvSimpson" if you prefer

# Choose ribbon type: "SE" or "CI95"
ribbon_type <- "CI95"

# Color palette (your existing)
group_colors <- c(
  "Vehicle"     = "#FFFFFF",
  "Haloperidol" = "#62666A",
  "Quetiapine"  = "#E9E4DC",
  "Olanzapine"  = "#81BDD1",
  "Risperidone" = "#C3AC84"
)

# ---- alpha diversity ----
richness_raw <- estimate_richness(ps, measures = c("Observed", "Shannon", "Simpson")) %>%
  rownames_to_column("SampleID") %>%
  mutate(InvSimpson = 1 / Simpson)

metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")

richness_df <- full_join(metadata_df, richness_raw, by = "SampleID") %>%
  filter(Time_Point %in% keep_time) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = keep_time),
    TreatmentGroup = factor(TreatmentGroup, levels = names(group_colors))
  ) %>%
  pivot_longer(
    cols = c("Observed", "Shannon", "InvSimpson"),
    names_to = "measure",
    values_to = "value"
  ) %>%
  filter(measure %in% c("Observed", evenness_metric))

# ---- summary stats for ribbons ----
summ_df <- richness_df %>%
  group_by(TreatmentGroup, Time_Point, measure) %>%
  summarise(
    n = sum(!is.na(value)),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    se = sd / sqrt(n),
    .groups = "drop"
  ) %>%
  mutate(
    # Use t critical value when n>1, else NA
    tcrit = ifelse(n > 1, qt(0.975, df = n - 1), NA_real_),
    ci_lo = mean - tcrit * se,
    ci_hi = mean + tcrit * se,
    se_lo = mean - se,
    se_hi = mean + se
  )

# Pick which bounds to use
summ_df <- summ_df %>%
  mutate(
    ymin = if (ribbon_type == "SE") se_lo else ci_lo,
    ymax = if (ribbon_type == "SE") se_hi else ci_hi
  )

# ---- plot ----
p <- ggplot(summ_df, aes(x = Time_Point, y = mean, group = TreatmentGroup)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax, fill = TreatmentGroup),
              alpha = 0.18, color = NA) +
  geom_line(aes(color = TreatmentGroup), linewidth = 1) +
  geom_point(aes(color = TreatmentGroup), size = 2) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_fill_manual(values = group_colors) +
  scale_color_manual(values = group_colors) +
  theme_classic(base_size = 12) +
  labs(
    title = paste0("Alpha Diversity Over Time (Mean ± ", ribbon_type, ")"),
    x = "Time Point",
    y = "Mean diversity"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

print(p)

# Save
ggsave(file.path(root.dir, paste0("alpha_diversity_lineplot_", evenness_metric, "_", ribbon_type, ".tiff")),
       plot = p, width = 14, height = 6, dpi = 300)
ggsave(file.path(root.dir, paste0("alpha_diversity_lineplot_", evenness_metric, "_", ribbon_type, ".pdf")),
       plot = p, width = 14, height = 6)

write.csv(summ_df,
          file = file.path(root.dir, paste0("alpha_diversity_summary_", evenness_metric, "_", ribbon_type, ".csv")),
          row.names = FALSE)

```
```{r}
install.packages("ggbeeswarm")
```
Beeswarm

```{r alpha_diversity_beeswarm_line, echo=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=7}
library(phyloseq)
library(tidyverse)
library(ggbeeswarm)

# --- Alpha diversity
richness_raw <- estimate_richness(ps, measures = c("Shannon", "Observed", "Simpson")) %>%
  rownames_to_column("SampleID")

metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")

richness_df <- left_join(metadata_df, richness_raw, by = "SampleID") %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6")),
    InverseSimpson = 1 / Simpson
  ) %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6"))

group_colors <- c(
  "Vehicle"     = "#000000",
  "Haloperidol" = "#62666A",
  "Quetiapine"  = "#E9E4DC",
  "Olanzapine"  = "#81BDD1",
  "Risperidone" = "#C3AC84"
)

# Force consistent order
richness_df$TreatmentGroup <- factor(richness_df$TreatmentGroup, levels = names(group_colors))

# Choose measures (swap Shannon -> InverseSimpson if desired)
measures_to_plot <- c("Observed", "Shannon")

plot_long <- richness_df %>%
  select(SampleID, Time_Point, TreatmentGroup, all_of(measures_to_plot)) %>%
  pivot_longer(cols = all_of(measures_to_plot), names_to = "measure", values_to = "value")

# Group means for overlay line
mean_df <- plot_long %>%
  group_by(TreatmentGroup, Time_Point, measure) %>%
  summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")

p <- ggplot() +
  # individual mice (beeswarm-like)
  geom_quasirandom(
    data = plot_long,
    aes(x = Time_Point, y = value, color = TreatmentGroup),
    dodge.width = 0.75,
    width = 0.18,
    alpha = 0.55,
    size = 1.6
  ) +
  # group mean trajectory
  geom_line(
    data = mean_df,
    aes(x = Time_Point, y = mean, group = TreatmentGroup, color = TreatmentGroup),
    linewidth = 1.2
  ) +
  geom_point(
    data = mean_df,
    aes(x = Time_Point, y = mean, color = TreatmentGroup),
    size = 2.4
  ) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_color_manual(values = group_colors) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Alpha diversity over time (individual mice + group mean)",
    x = "Time point",
    y = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

print(p)

ggsave(file.path(root.dir, "alpha_diversity_beeswarm_meanline.tiff"),
       p, width = 16, height = 7, dpi = 300)
ggsave(file.path(root.dir, "alpha_diversity_beeswarm_meanline.pdf"),
       p, width = 16, height = 7)
```

```{r alpha_diversity_beeswarm_hollow_vehicle, echo=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=7}
library(phyloseq)
library(tidyverse)
library(ggbeeswarm)

# ---- Alpha diversity
richness_raw <- estimate_richness(ps, measures = c("Shannon", "Observed", "Simpson")) %>%
  rownames_to_column("SampleID")

metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")

richness_df <- left_join(metadata_df, richness_raw, by = "SampleID") %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6")),
    InverseSimpson = 1 / Simpson
  ) %>%
  filter(Time_Point %in% c("T0","T2","T3","T6"))

group_colors <- c(
  "Vehicle"     = "#000000",   # outline only
  "Haloperidol" = "#62666A",
  "Quetiapine"  = "#E9E4DC",
  "Olanzapine"  = "#81BDD1",
  "Risperidone" = "#C3AC84"
)

richness_df$TreatmentGroup <- factor(richness_df$TreatmentGroup, levels = names(group_colors))

# Choose measures
measures_to_plot <- c("Observed", "Shannon")  # or c("Observed", "InverseSimpson")

plot_long <- richness_df %>%
  select(SampleID, Time_Point, TreatmentGroup, all_of(measures_to_plot)) %>%
  pivot_longer(cols = all_of(measures_to_plot), names_to = "measure", values_to = "value")

# Group means
mean_df <- plot_long %>%
  group_by(TreatmentGroup, Time_Point, measure) %>%
  summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")

# Flag Vehicle for hollow points
plot_long <- plot_long %>%
  mutate(
    fill_group = if_else(TreatmentGroup == "Vehicle", NA_character_, as.character(TreatmentGroup))
  )

p <- ggplot() +
  # Individual mice
  geom_quasirandom(
    data = plot_long,
    aes(
      x = Time_Point,
      y = value,
      color = TreatmentGroup,
      fill  = fill_group
    ),
    shape = 21,                 # allows fill + outline
    stroke = 0.8,
    dodge.width = 0.75,
    width = 0.18,
    alpha = 0.6,
    size = 1.8
  ) +
  # Mean trajectory
  geom_line(
    data = mean_df,
    aes(x = Time_Point, y = mean, group = TreatmentGroup, color = TreatmentGroup),
    linewidth = 1.2
  ) +
  geom_point(
    data = mean_df,
    aes(x = Time_Point, y = mean, color = TreatmentGroup),
    size = 2.6
  ) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_color_manual(values = group_colors) +
  scale_fill_manual(values = group_colors, na.value = NA) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Alpha diversity over time (individual mice + group mean)",
    x = "Time point",
    y = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

print(p)

ggsave(file.path(root.dir, "alpha_diversity_beeswarm_hollow_vehicle.tiff"),
       p, width = 16, height = 7, dpi = 300)
ggsave(file.path(root.dir, "alpha_diversity_beeswarm_hollow_vehicle.pdf"),
       p, width = 16, height = 7)
```

2. Alpha Diversity (across Groups and Time)
```{r alpha_diversity_plot, echo=FALSE, message=FALSE, warning=FALSE, fig.width=16, fig.height=8}

# Prepare richness data including Simpson
richness_raw <- estimate_richness(ps, measures = c("Shannon", "Observed", "Simpson")) %>%
  rownames_to_column("SampleID")

# Load metadata
metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")

# Merge and reshape
richness_df <- full_join(metadata_df, richness_raw, by = "SampleID") %>%
  pivot_longer(cols = c("Shannon", "Observed", "Simpson"), names_to = "measure", values_to = "value")

# Filter and re-level time points
richness_df <- richness_df %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6"))

richness_df$Time_Point <- factor(richness_df$Time_Point, levels = c("T0", "T2", "T3", "T6"))
richness_df$TreatmentGroup <- factor(richness_df$TreatmentGroup, levels = unique(metadata_df$TreatmentGroup))

# Generate the plot
p <- ggplot(richness_df, aes(x = Time_Point, y = value, fill = TreatmentGroup)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, position = position_dodge(width = 0.8)) +
  geom_jitter(
    aes(color = TreatmentGroup),
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8),
    size = 1,
    alpha = 0.5
  ) +
  geom_smooth(
    aes(group = TreatmentGroup, color = TreatmentGroup),
    method = "lm",
    se = FALSE,
    linetype = "dashed"
  ) +
  facet_wrap(~ measure, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Alpha Diversity Over Time by Treatment Group",
    x = "Time Point",
    y = "Diversity Index"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right"
  )

# Display the plot
print(p)

# Save the plot
ggsave(filename = file.path(root.dir, "alpha_diversity_plot.tiff"), plot = p, width = 16, height = 8, dpi = 300)
ggsave(filename = file.path(root.dir, "alpha_diversity_plot.pdf"), plot = p, width = 16, height = 8)



```

```{r alpha_diversity_faceted_hollow_vehicle_shannon, echo=FALSE, message=FALSE, warning=FALSE, fig.width=18, fig.height=8}
library(phyloseq)
library(tidyverse)
library(ggbeeswarm)

# ---- Alpha diversity
richness_raw <- estimate_richness(ps, measures = c("Observed", "Shannon")) %>%
  rownames_to_column("SampleID")

metadata_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")

richness_df <- left_join(metadata_df, richness_raw, by = "SampleID") %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6"))
  ) %>%
  filter(Time_Point %in% c("T0","T2","T3","T6"))

# ---- Original group colors
group_colors <- c(
  "Vehicle"     = "#FFFFFF",   # fill for points (vehicle will be white)
  "Haloperidol" = "#62666A",
  "Quetiapine"  = "#E9E4DC",
  "Olanzapine"  = "#81BDD1",
  "Risperidone" = "#C3AC84"
)

# Mean line colors: Vehicle forced to black
line_colors <- c(
  "Vehicle"     = "#000000",
  "Haloperidol" = "#62666A",
  "Quetiapine"  = "#E9E4DC",
  "Olanzapine"  = "#81BDD1",
  "Risperidone" = "#C3AC84"
)

richness_df$TreatmentGroup <- factor(richness_df$TreatmentGroup, levels = names(group_colors))

# ---- Long format: Observed + Shannon only
plot_long <- richness_df %>%
  select(SampleID, Time_Point, TreatmentGroup, Observed, Shannon) %>%
  pivot_longer(cols = c(Observed, Shannon),
               names_to = "measure", values_to = "value") %>%
  mutate(
    measure = factor(
      measure,
      levels = c("Observed", "Shannon"),
      labels = c("Observed richness", "Shannon diversity")
    ),
    # Vehicle dots: white fill; other groups: colored fill
    fill_group = if_else(TreatmentGroup == "Vehicle", NA_character_, as.character(TreatmentGroup))
  )

# ---- Group means
mean_df <- plot_long %>%
  group_by(TreatmentGroup, Time_Point, measure) %>%
  summarise(mean = mean(value, na.rm = TRUE), .groups = "drop")

p <- ggplot() +
  # Individual mice (beeswarm) with black outline for ALL dots
  geom_quasirandom(
    data = plot_long,
    aes(x = Time_Point, y = value, fill = fill_group),
    shape = 21,
    color = "black",     # outline
    stroke = 0.7,
    width = 0.18,
    alpha = 0.65,
    size = 1.8
  ) +
  # Mean trajectory (Vehicle line in black)
  geom_line(
    data = mean_df,
    aes(x = Time_Point, y = mean, group = TreatmentGroup, color = TreatmentGroup),
    linewidth = 1.2
  ) +
  geom_point(
    data = mean_df,
    aes(x = Time_Point, y = mean, color = TreatmentGroup),
    size = 2.6
  ) +
  facet_grid(measure ~ TreatmentGroup, scales = "free_y") +
  scale_fill_manual(values = group_colors, na.value = "white") +
  scale_color_manual(values = line_colors) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Alpha diversity over time by treatment_shannon",
    x = "Time point",
    y = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
print(p)

ggsave(filename = file.path(root.dir, "alpha_diversity_plot_shannon.tiff"), plot = p, width = 16, height = 8, dpi = 300)
ggsave(filename = file.path(root.dir, "alpha_diversity_plot_shannon.pdf"), plot = p, width = 16, height = 8)
```


### Explanation:
- `facet_grid(measure ~ TreatmentGroup)` splits the plot into rows by diversity measure (`Shannon`, `Observed`) and columns by treatment group.
- Lines connect samples over time (`SampleID` groups points).
- Points are colored by `SampleID` to track individuals (legend hidden for clarity).
- Scales are free per facet to accommodate different diversity ranges.

If you prefer **summary lines (mean ± error)** instead of individual trajectories, I can help with that too!


```{r alpha_diversity_plot, echo=FALSE, warning=FALSE, message=FALSE, fig.width=15, fig.height=10}
library(ggplot2)
library(dplyr)
library(broom)
library(tidyr)

# Step 1: Add numeric time for linear regression
richness_df <- richness_df %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6")) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6")),
    Time_Point_num = as.numeric(Time_Point)
  )

# Step 2: Fit linear models per TreatmentGroup and measure
lm_stats <- richness_df %>%
  group_by(TreatmentGroup, measure) %>%
  do({
    model <- lm(value ~ Time_Point_num, data = .)
    tidy_mod <- tidy(model)
    glance_mod <- glance(model)
    data.frame(
      intercept = tidy_mod$estimate[1],
      slope = tidy_mod$estimate[2],
      r_squared = glance_mod$r.squared
    )
  }) %>%
  ungroup() %>%
  mutate(
    eq_label = paste0("y = ", round(slope, 3), "x + ", round(intercept, 3), 
                      "; R² = ", round(r_squared, 3))
  )

# Step 3: Determine annotation position
x_positions <- richness_df %>%
  group_by(TreatmentGroup, measure) %>%
  summarise(
    x = mean(Time_Point_num, na.rm = TRUE),
    .groups = "drop"
  )

y_positions <- richness_df %>%
  group_by(measure, TreatmentGroup) %>%
  summarise(
    y = min(value, na.rm = TRUE) - 0.1 * (max(value, na.rm = TRUE) - min(value, na.rm = TRUE)),
    .groups = "drop"
  )

lm_stats <- lm_stats %>%
  left_join(y_positions, by = c("measure", "TreatmentGroup")) %>%
  left_join(x_positions, by = c("measure", "TreatmentGroup"))

# Step 4: Plot
p <- ggplot(richness_df, aes(x = Time_Point_num, y = value, group = SampleID, color = SampleID)) +
  geom_point(alpha = 0.6) +
  geom_line(alpha = 0.6) +
  geom_smooth(aes(group = TreatmentGroup), method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  facet_grid(measure ~ TreatmentGroup, scales = "free_y", switch = "y") +
  scale_x_continuous(
    breaks = 1:4,
    labels = c("T0", "T2", "T3", "T6"),
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Alpha Diversity Over Time by Treatment Group with Linear Regression",
    x = "Time Point",
    y = "Diversity Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.margin = margin(10, 10, 60, 10),
    panel.spacing.x = unit(1, "cm"),
    panel.spacing.y = unit(1, "cm"),
    axis.text.x = element_text(margin = margin(t = 10))
  ) +
  geom_text(
    data = lm_stats,
    aes(x = x, y = y, label = eq_label),
    inherit.aes = FALSE,
    size = 2.8,
    hjust = 0.5,
    vjust = 0,
    fontface = "italic"
  )

print(p)

# Step 5: Save
ggsave(filename = file.path(root.dir, "alpha_diversity_all_metrics.pdf"), plot = p, width = 15, height = 10)
ggsave(filename = file.path(root.dir, "alpha_diversity_all_metrics.tiff"), plot = p, width = 15, height = 10, dpi = 300)

```



```{r}
# Export the diversity data used for plotting
write.csv(richness_df, file = file.path(root.dir, "alpha_diversity_data.csv"), row.names = FALSE)

```

Equation Exports
```{r diversity equations, fig.width=100, fig.height=0.5}
richness_df <- richness_df %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6")) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6")),
    Time_Point_num = as.numeric(Time_Point)
  )

# Define directory inside root.dir for saving equation plots
eq_dir <- file.path(root.dir, "lm_equations")

# Add these columns first
# Add columns for equation and R2 label
lm_stats <- lm_stats %>%
  mutate(
    equation = paste0("y = ", round(slope, 3), "x + ", round(intercept, 3)),
    r2_label = paste0("R² = ", round(r_squared, 3))
  )

# Define directory for equation plots
eq_dir <- file.path(root.dir, "lm_equations")
if (!dir.exists(eq_dir)) {
  dir.create(eq_dir, recursive = TRUE)
}

# Export CSV summary
lm_export <- lm_stats %>%
  dplyr::select(TreatmentGroup, measure, equation, r2_label) %>%
  dplyr::rename(
    Method = measure,
    Equation = equation,
    R_squared = r2_label
  )


write.csv(lm_export, file = file.path(eq_dir, "lm_equations_summary.csv"), row.names = FALSE)

# Loop to save plots
for(i in seq_len(nrow(lm_stats))) {
  row <- lm_stats[i, ]
  
  p <- ggplot() +
    geom_text(
      aes(x = 0.5, y = 0.7, label = equation),
      data = row,
      size = 6,
      hjust = 0.5, vjust = 1
    ) +
    geom_text(
      aes(x = 0.5, y = 0.2, label = r2_label),
      data = row,
      size = 6,
      hjust = 0.5, vjust = 1
    ) +
    theme_void() +
    theme(plot.margin = margin(5, 5, 5, 5)) +
    xlim(0, 1) + ylim(-0.4, 1) +
    coord_cartesian(clip = "off")
  
  base_fname <- paste0("lm_eq_", row$measure, "_", row$TreatmentGroup)
  base_fname <- gsub(" ", "_", base_fname)
  
  tiff_path <- file.path(eq_dir, paste0(base_fname, ".tiff"))
  pdf_path <- file.path(eq_dir, paste0(base_fname, ".pdf"))
  
  ggsave(filename = tiff_path, plot = p, width = 3, height = 1, dpi = 300, device = "tiff")
  ggsave(filename = pdf_path, plot = p, width = 3, height = 1, device = "pdf")
}




```

```{r regression_statistics, message=FALSE, warning=FALSE}
library(dplyr)
library(broom)
library(purrr)

# Filter data to relevant time points and ensure factors are correctly set
richness_df <- richness_df %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6")) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6")),
    Time_Point_num = as.numeric(Time_Point),
    TreatmentGroup = factor(TreatmentGroup, levels = c("Vehicle", setdiff(levels(factor(TreatmentGroup)), "Vehicle")))
  )


# Function to fit linear model with interaction term and extract interaction stats
compare_slopes_vs_vehicle <- function(df) {
  model <- lm(value ~ Time_Point_num * TreatmentGroup, data = df)
  tidy_mod <- broom::tidy(model)
  
  # Extract only interaction terms (Time_Point_num:TreatmentGroup)
  interaction_terms <- tidy_mod %>%
    filter(grepl("Time_Point_num:TreatmentGroup", term)) %>%
    dplyr::select(term, estimate, std.error, statistic, p.value)
  
  # Return empty tibble with expected columns if no interactions found
  if (nrow(interaction_terms) == 0) {
    return(tibble(
      term = character(0),
      estimate = numeric(0),
      std.error = numeric(0),
      statistic = numeric(0),
      p.value = numeric(0)
    ))
  }
  
  return(interaction_terms)
}

# Apply the comparison function to each 'measure' group separately
comparison_results <- richness_df %>%
  group_by(measure) %>%
  group_split() %>%
  map_df(~ {
    df <- .
    res <- compare_slopes_vs_vehicle(df)
    
    # Add 'measure' and cleaned-up group names for clarity
    res <- res %>%
      mutate(
        measure = unique(df$measure),
        ComparedGroup = gsub("Time_Point_num:TreatmentGroup", "", term)
      ) %>%
      dplyr::select(measure, ComparedGroup, estimate, std.error, statistic, p.value)
    res
  })

# Print the results: p-values here indicate if the slope differs from Vehicle's slope
print(comparison_results)
# Save the results to a CSV file
write.csv(comparison_results, file = file.path(root.dir, "slope_comparison_results.csv"), row.names = FALSE)

```

```{r alpha_diversity_with_pvalues, message=FALSE, warning=FALSE,  fig.width=15, fig.height=10}
library(ggplot2)
library(dplyr)
library(broom)
library(purrr)
library(gridExtra)   # for tableGrob
library(grid)        # for grid.arrange
library(patchwork)   # for plot + table layout

# Filter and set factor levels with Vehicle as reference
richness_df <- richness_df %>%
  filter(Time_Point %in% c("T0", "T2", "T3", "T6")) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = c("T0", "T2", "T3", "T6")),
    Time_Point_num = as.numeric(Time_Point),
    TreatmentGroup = factor(TreatmentGroup, levels = c("Vehicle", setdiff(unique(TreatmentGroup), "Vehicle")))
  )

# Plot: Alpha diversity over time by treatment, faceted by measure and TreatmentGroup
p <- ggplot(richness_df, aes(x = Time_Point_num, y = value, group = SampleID, color = TreatmentGroup)) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  geom_smooth(aes(group = TreatmentGroup), method = "lm", se = FALSE, linetype = "dashed", color = "black") +
  facet_grid(measure ~ TreatmentGroup, scales = "free_y") +
  scale_x_continuous(
    breaks = 1:4,
    labels = c("T0", "T2", "T3", "T6"),
    expand = expansion(mult = c(0.05, 0.1))
  ) +
  labs(
    title = "Alpha Diversity Over Time by Treatment Group",
    x = "Time Point",
    y = "Diversity Index"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

# Function to extract slope interaction p-values comparing groups vs Vehicle
compare_slopes_vs_vehicle <- function(df) {
  model <- lm(value ~ Time_Point_num * TreatmentGroup, data = df)
  tidy_mod <- broom::tidy(model)
  
  interaction_terms <- tidy_mod %>%
    filter(grepl("Time_Point_num:TreatmentGroup", term)) %>%
    dplyr::select(term, estimate, std.error, statistic, p.value)
  
  if (nrow(interaction_terms) == 0) {
    return(tibble(
      term = character(0),
      estimate = numeric(0),
      std.error = numeric(0),
      statistic = numeric(0),
      p.value = numeric(0)
    ))
  }
  return(interaction_terms)
}

# Get comparison results for each measure
comparison_results <- richness_df %>%
  group_by(measure) %>%
  group_split() %>%
  map_df(~ {
    df <- .
    res <- compare_slopes_vs_vehicle(df)
    res %>%
      mutate(
        measure = unique(df$measure),
        ComparedGroup = gsub("Time_Point_num:TreatmentGroup", "", term),
        p.value = signif(p.value, 3)
      ) %>%
      dplyr::select(measure, ComparedGroup, estimate, std.error, statistic, p.value)
  })

# Format p-value table for display
pval_table <- comparison_results %>%
  mutate(
    ComparedGroup = factor(ComparedGroup, levels = levels(richness_df$TreatmentGroup)),
    measure = factor(measure)
  ) %>%
  arrange(measure, ComparedGroup) %>%
  dplyr::rename(
    Measure = measure,
    Group = ComparedGroup,
    `Slope Estimate` = estimate,
    `p-value` = p.value
  ) %>%
  dplyr::select(Measure, Group, `Slope Estimate`, `p-value`)

# Create a grob table
table_grob <- gridExtra::tableGrob(pval_table, rows = NULL)

# Use patchwork to stack plot and table; wrap_elements needed to embed grob
final_plot <- p / patchwork::wrap_elements(table_grob) + plot_layout(heights = c(2, 2))

# Print combined figure
print(final_plot)

# Save combined plot (optional)
ggsave(filename = file.path(root.dir, "alpha_diversity_with_pvalues.pdf"), plot = final_plot, width = 15, height = 10)
ggsave(filename = file.path(root.dir, "alpha_diversity_with_pvalues.tiff"), plot = final_plot, width = 15, height = 10, dpi = 300)


```

3. Beta Diversity with PCoa (Bray or UniFrac)
```{r}
library(phyloseq)
library(ggplot2)

# Rarefy only for ordination (NOT for DESeq2 or richness)
ps.rarefied <- rarefy_even_depth(ps, rngseed = 42, verbose = FALSE)

# Subset rarefied phyloseq object to keep only T0, T2, T3, T6
ps.sub <- subset_samples(ps.rarefied, Time_Point %in% c("T0", "T2", "T3", "T6"))

# Perform ordination on filtered samples
ord <- ordinate(ps.sub, method = "PCoA", distance = "bray")

# Plot with filtered samples
p <- plot_ordination(ps.sub, ord, color = "TreatmentGroup", shape = "Time_Point") +
  theme_minimal() +
  labs(title = "Bray-Curtis PCoA (T0, T2, T3, T6)", color = "Treatment", shape = "Time Point")

print(p)  # To display inline in RMarkdown or console

# Save plot as PDF and TIFF in root directory
ggsave(filename = file.path(root.dir, "Bray_Curtis_PCoA_T0_T2_T3_T6.pdf"), plot = p, device = "pdf", width = 8, height = 6)
ggsave(filename = file.path(root.dir, "Bray_Curtis_PCoA_T0_T2_T3_T6.tiff"), plot = p, device = "tiff", dpi = 300, width = 8, height = 6)

```
Beta Diversity for each time point
```{r pcoa_plots, fig.width=15, fig.height=10}
library(phyloseq)
library(ggplot2)
library(patchwork)
library(ggforce)  # for geom_mark_ellipse()

# Define only the time points you want to plot
time_points <- c("T0", "T2", "T3", "T6")

pcoas <- vector("list", length(time_points))

for (i in seq_along(time_points)) {
  tp <- time_points[i]
  
  # Subset rarefied phyloseq object by time point
  ps_sub <- subset_samples(ps.rarefied, Time_Point == tp)
  
  # Remove taxa with zero counts after subsetting
  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
  
  # Perform ordination
  ord_sub <- ordinate(ps_sub, method = "PCoA", distance = "bray")
  
  # Extract ordination data for plotting
  ord_df <- plot_ordination(ps_sub, ord_sub, color = "TreatmentGroup")$data
  
  # Calculate buffered axis limits
  x_range <- range(ord_df$Axis.1)
  x_buffer <- diff(x_range) * 0.10
  x_limits <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
  
  y_range <- range(ord_df$Axis.2)
  y_buffer <- diff(y_range) * 0.10
  y_limits <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)
  
  # Create PCoA plot for this time point
  pcoas[[i]] <- ggplot(ord_df, aes(x = Axis.1, y = Axis.2, color = TreatmentGroup)) +
    geom_point(size = 3, alpha = 0.7) +
    ggforce::geom_mark_ellipse(aes(group = TreatmentGroup, fill = TreatmentGroup), alpha = 0.1, show.legend = FALSE) +
    labs(title = tp, x = "PCoA1", y = "PCoA2") +
    theme_minimal() +
    theme(
      legend.position = ifelse(i == length(time_points), "right", "none"),
      plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm")
    ) +
    coord_cartesian(xlim = x_limits, ylim = y_limits, clip = "off")
}

# Combine and print the plots
combined_plot <- wrap_plots(pcoas, ncol = 2, nrow = 2) + 
  plot_layout(guides = "collect") & 
  theme(plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))

print(combined_plot)

# Save combined plot to root directory
ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_Filtered_TimePoints.pdf"),
  plot = combined_plot,
  width = 15,
  height = 10,
  device = "pdf"
)

ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_Filtered_TimePoints.tiff"),
  plot = combined_plot,
  width = 15,
  height = 10,
  dpi = 300,
  device = "tiff"
)



```


```{r pcoa_plots, fig.width=15, fig.height=10}
library(phyloseq)
library(ggplot2)
library(patchwork)
library(ggforce)  # for geom_mark_ellipse()

# Define only the time points you want to plot
time_points <- c("T0", "T2", "T3", "T6")

# Define consistent color palette (same as alpha diversity)
group_colors <- c(
  "Vehicle" = "#1b9e77",
  "Haloperidol" = "#d95f02",
  "Quetiapine" = "#7570b3",
  "Risperidone" = "#e7298a",
  "Olanzapine" = "#66a61e"
)

pcoas <- vector("list", length(time_points))

for (i in seq_along(time_points)) {
  tp <- time_points[i]
  
  # Subset rarefied phyloseq object by time point
  ps_sub <- subset_samples(ps.rarefied, Time_Point == tp)
  
  # Remove taxa with zero counts after subsetting
  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
  
  # Perform ordination
  ord_sub <- ordinate(ps_sub, method = "PCoA", distance = "bray")
  
  # Extract ordination data for plotting
  ord_df <- plot_ordination(ps_sub, ord_sub, color = "TreatmentGroup")$data
  
  # Ensure TreatmentGroup is a factor with the same order as before
  ord_df$TreatmentGroup <- factor(ord_df$TreatmentGroup, 
                                  levels = names(group_colors))
  
  # Calculate buffered axis limits
  x_range <- range(ord_df$Axis.1)
  x_buffer <- diff(x_range) * 0.10
  x_limits <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
  
  y_range <- range(ord_df$Axis.2)
  y_buffer <- diff(y_range) * 0.10
  y_limits <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)
  
  # Create PCoA plot for this time point
  pcoas[[i]] <- ggplot(ord_df, aes(x = Axis.1, y = Axis.2, color = TreatmentGroup)) +
    geom_point(size = 3, alpha = 0.7) +
    ggforce::geom_mark_ellipse(aes(group = TreatmentGroup, fill = TreatmentGroup),
                               alpha = 0.1, show.legend = FALSE) +
    scale_color_manual(values = group_colors) +
    scale_fill_manual(values = group_colors) +
    labs(title = tp, x = "PCoA1", y = "PCoA2") +
    theme_minimal() +
    theme(
      legend.position = ifelse(i == length(time_points), "right", "none"),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")
    ) +
    coord_cartesian(xlim = x_limits, ylim = y_limits, clip = "off")
}

# Combine and print the plots
combined_plot <- wrap_plots(pcoas, ncol = 2, nrow = 2) + 
  plot_layout(guides = "collect") & 
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

print(combined_plot)

# Save combined plot to root directory
ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_Filtered_TimePoints_2.pdf"),
  plot = combined_plot,
  width = 15,
  height = 10,
  device = "pdf"
)

ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_Filtered_TimePoints_2.tiff"),
  plot = combined_plot,
  width = 15,
  height = 10,
  dpi = 300,
  device = "tiff"
)
```

PCOA Plots per Treatment group instead of by time point 

```{r pcoa_plots, fig.width=15, fig.height=10}
library(phyloseq)
library(ggplot2)
library(patchwork)
library(ggforce)
library(dplyr)

# Define only the time points you want to include
time_points <- c("T0", "T2", "T3", "T6")

# Remove unwanted time points (e.g., T8)
ps_filtered <- subset_samples(ps.rarefied, Time_Point %in% time_points)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# Define colors for Time Points
time_colors <- c(
  "T0" = "#1b9e77",
  "T2" = "#d95f02",
  "T3" = "#7570b3",
  "T6" = "#e7298a"
)

# Treatment groups
treatments <- c("Vehicle", "Haloperidol", "Quetiapine", "Risperidone", "Olanzapine")

pcoas <- vector("list", length(treatments))

for (i in seq_along(treatments)) {
  trt <- treatments[i]
  
  # Subset by treatment
  ps_sub <- subset_samples(ps_filtered, TreatmentGroup == trt)
  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)
  
  # Perform PCoA (Bray-Curtis)
  ord_sub <- ordinate(ps_sub, method = "PCoA", distance = "bray")
  
  # Extract ordination data for plotting
  ord_df <- plot_ordination(ps_sub, ord_sub, color = "Time_Point")$data %>%
    filter(!is.na(Time_Point)) %>%  # ensure no NA group
    mutate(Time_Point = factor(Time_Point, levels = time_points))
  
  # Calculate buffered axis limits
  x_range <- range(ord_df$Axis.1)
  x_buffer <- diff(x_range) * 0.10
  x_limits <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
  
  y_range <- range(ord_df$Axis.2)
  y_buffer <- diff(y_range) * 0.10
  y_limits <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)
  
  # Create PCoA plot for this treatment
  pcoas[[i]] <- ggplot(ord_df, aes(x = Axis.1, y = Axis.2, color = Time_Point)) +
    geom_point(size = 3, alpha = 0.8) +
    ggforce::geom_mark_ellipse(aes(group = Time_Point, fill = Time_Point),
                               alpha = 0.1, show.legend = FALSE) +
    scale_color_manual(values = time_colors) +
    scale_fill_manual(values = time_colors) +
    labs(title = trt, x = "PCoA1", y = "PCoA2") +
    theme_minimal() +
    theme(
      legend.position = ifelse(i == length(treatments), "right", "none"),
      plot.title = element_text(face = "bold", size = 14),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")
    ) +
    coord_cartesian(xlim = x_limits, ylim = y_limits, clip = "off")
}

# Combine and print plots
combined_plot <- wrap_plots(pcoas, ncol = 3, nrow = 2) +
  plot_layout(guides = "collect") &
  theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm"))

print(combined_plot)

# Save combined plots
ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_ByTreatment_T0T6.pdf"),
  plot = combined_plot,
  width = 15,
  height = 10,
  device = "pdf"
)

ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_ByTreatment_T0T6.tiff"),
  plot = combined_plot,
  width = 15,
  height = 10,
  dpi = 300,
  device = "tiff"
)

```

single-ordination + facet-by-treatment

```{r pcoa_faceted_single_ordination, fig.width=15, fig.height=10}
library(phyloseq)
library(ggplot2)
library(ggforce)
library(dplyr)

# ---- Settings ----
time_points <- c("T0", "T2", "T3", "T6")  # include only these time points

time_colors <- c(
  "T0" = "#1b9e77",
  "T2" = "#d95f02",
  "T3" = "#7570b3",
  "T6" = "#e7298a"
)

# Keep a consistent order for facet panels (optional)
treatments <- c("Vehicle", "Haloperidol", "Quetiapine", "Risperidone", "Olanzapine")

# ---- Filter to desired time points and prune zero-count taxa ----
ps_filtered <- subset_samples(ps.rarefied, Time_Point %in% time_points)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# ---- Single ordination on the entire filtered dataset ----
ord_full <- ordinate(ps_filtered, method = "PCoA", distance = "bray")

# Extract ordination data merged with metadata
ord_df <- plot_ordination(ps_filtered, ord_full, color = "Time_Point", type = "samples")$data %>%
  filter(!is.na(Time_Point)) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = time_points),
    TreatmentGroup = factor(TreatmentGroup, levels = treatments)
  )

# Optional: compute % variance explained for axis labels (if available)
axis_labels <- c("PCoA1", "PCoA2")
if (!is.null(ord_full$values) && !is.null(ord_full$values$Relative_eig)) {
  var_exp <- round(100 * ord_full$values$Relative_eig[1:2], 1)
  axis_labels <- c(
    paste0("PCoA1 (", var_exp[1], "%)"),
    paste0("PCoA2 (", var_exp[2], "%)")
  )
}

# ---- Plot: single ordination, faceted by TreatmentGroup ----
p_faceted <- ggplot(ord_df, aes(x = Axis.1, y = Axis.2, color = Time_Point)) +
  geom_point(size = 3, alpha = 0.85) +
  ggforce::geom_mark_ellipse(
    aes(group = Time_Point, fill = Time_Point),
    alpha = 0.12, show.legend = FALSE
  ) +
  scale_color_manual(values = time_colors, drop = FALSE) +
  scale_fill_manual(values = time_colors, drop = FALSE) +
  facet_wrap(~ TreatmentGroup, ncol = 3, drop = FALSE) +
  coord_fixed() +  # same units on x and y for comparability
  labs(title = "PCoA (Bray–Curtis) — Faceted by Treatment",
       x = axis_labels[1], y = axis_labels[2],
       color = "Time Point") +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

print(p_faceted)

# ---- Save ----
ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_SingleOrdination_FacetedByTreatment.pdf"),
  plot = p_faceted, width = 15, height = 10, device = "pdf"
)
ggsave(
  filename = file.path(root.dir, "PCoA_Bray_Curtis_SingleOrdination_FacetedByTreatment.tiff"),
  plot = p_faceted, width = 15, height = 10, dpi = 300, device = "tiff"
)
```
Bray Curtis dot plot medoid

```{r bray_to_T0_dotplot_full, fig.width=14, fig.height=9}
library(phyloseq)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(vegan)

# ---- Settings ----
time_points <- c("T0", "T2", "T3", "T6")
treatments  <- c("Vehicle", "Haloperidol", "Quetiapine", "Risperidone", "Olanzapine")

time_colors <- c("T0"="#1b9e77","T2"="#d95f02","T3"="#7570b3","T6"="#e7298a")

# ---- Filter & prune ----
ps_filtered <- subset_samples(ps.rarefied, Time_Point %in% time_points)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# ---- metadata and matrix (samples as rows) ----
meta <- data.frame(sample_data(ps_filtered), stringsAsFactors = FALSE)
meta$SampleID <- rownames(meta)

mat <- as(otu_table(ps_filtered), "matrix")
if (taxa_are_rows(ps_filtered)) mat <- t(mat)
rel <- sweep(mat, 1, pmax(rowSums(mat), 1e-12), "/")

# ---- T0 medoid fallback per treatment ----
get_medoid_id <- function(sample_ids) {
  if (length(sample_ids) <= 1) return(ifelse(length(sample_ids)==1, sample_ids, NA_character_))
  dm <- vegdist(rel[sample_ids, , drop = FALSE], method = "bray")
  m  <- as.matrix(dm)
  medoid_idx <- which.min(rowMeans(m))
  sample_ids[medoid_idx]
}

t0_meta <- meta %>% filter(Time_Point == "T0")
treat_medoid <- t0_meta %>%
  group_by(TreatmentGroup) %>%
  summarize(MedoidID = get_medoid_id(SampleID), .groups = "drop")
global_medoid <- get_medoid_id(t0_meta$SampleID)

# ---- Map each sample to baseline (per treatment x mouse) ----
mouse_t0_map <- t0_meta %>%
  group_by(TreatmentGroup, Mouse_Number) %>%  # <- key change
  slice(1) %>%
  ungroup() %>%
  select(TreatmentGroup, Mouse_Number, T0ID = SampleID)

meta <- meta %>%
  left_join(mouse_t0_map, by = c("TreatmentGroup", "Mouse_Number")) %>%
  left_join(treat_medoid, by = "TreatmentGroup") %>%
  mutate(
    BaselineID = dplyr::coalesce(T0ID, MedoidID, global_medoid),
    # and if you want all T0 samples at distance 0:
    BaselineID = dplyr::if_else(Time_Point == "T0", SampleID, BaselineID)
  )


# ---- Compute Bray–Curtis vs baseline (T0 included as 0) ----
compute_dist_to_baseline <- function(sid, bid) {
  if (is.na(bid) || is.na(sid)) return(NA_real_)
  if (sid == bid) return(0)  # include T0 as 0
  as.numeric(vegdist(rbind(rel[bid, , drop = FALSE],
                           rel[sid, , drop = FALSE]), method = "bray"))[1]
}

dist_df <- meta %>%
  mutate(Bray_to_Baseline = map2_dbl(SampleID, BaselineID, compute_dist_to_baseline)) %>%
  mutate(
    Time_Point = factor(Time_Point, levels = time_points),
    TreatmentGroup = factor(TreatmentGroup, levels = treatments)
  )

# ---- Numeric x for LOESS ----
dist_df <- dist_df %>%
  mutate(tp_num = as.integer(factor(Time_Point, levels = c("T0","T2","T3","T6"))))

# ---- Plot ----
p <- ggplot(dist_df, aes(x = tp_num, y = Bray_to_Baseline, color = Time_Point)) +
  geom_point(position = position_jitter(width = 0.08, height = 0),
             alpha = 0.75, size = 2.5) +
  geom_smooth(aes(group = 1),
              method = "loess", se = TRUE, span = 0.9,
              linewidth = 1.0, color = "black") +
  scale_x_continuous(breaks = 1:4,
                     labels = c("T0","T2","T3","T6"),
                     limits = c(1,4)) +
  scale_color_manual(values = time_colors, drop = FALSE) +
  facet_wrap(~ TreatmentGroup, ncol = 3, drop = FALSE) +
  labs(title = "Bray–Curtis Dissimilarity vs T0 Baseline",
       x = "Time Point", y = "Bray–Curtis vs Baseline") +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold", size = 16),
        legend.position = "none",
        panel.grid.minor = element_blank())

print(p)

# ---- Save ----
ggsave(file.path(root.dir,
        "BrayCurtis_vs_T0_Dotplot_byTreatment_includesT0.pdf"),
       p, width = 14, height = 9, device = "pdf")
ggsave(file.path(root.dir,
        "BrayCurtis_vs_T0_Dotplot_byTreatment_includesT0.tiff"),
       p, width = 14, height = 9, dpi = 300, device = "tiff")


```

The global T0 medoid is the T0 sample whose microbial profile is
closest to all the other T0 samples on average
```{r bray_to_globalT0_dotplot_full, fig.width=14, fig.height=9}
library(phyloseq)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(vegan)

# ---- Settings ----
time_points <- c("T0", "T2", "T3", "T6")
treatments  <- c("Vehicle", "Haloperidol", "Quetiapine", "Risperidone", "Olanzapine")

time_colors <- c("T0"="#1b9e77","T2"="#d95f02","T3"="#7570b3","T6"="#e7298a")

# ---- Filter & prune ----
ps_filtered <- subset_samples(ps.rarefied, Time_Point %in% time_points)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# ---- metadata and matrix (samples as rows) ----
meta <- data.frame(sample_data(ps_filtered), stringsAsFactors = FALSE)
meta$SampleID <- rownames(meta)

mat <- as(otu_table(ps_filtered), "matrix")
if (taxa_are_rows(ps_filtered)) mat <- t(mat)

# Relative abundance per sample
rel <- sweep(mat, 1, pmax(rowSums(mat), 1e-12), "/")

# ---- Helper: T0 global medoid ----
get_medoid_id <- function(sample_ids) {
  if (length(sample_ids) <= 1) {
    return(ifelse(length(sample_ids) == 1, sample_ids, NA_character_))
  }
  dm <- vegdist(rel[sample_ids, , drop = FALSE], method = "bray")
  m  <- as.matrix(dm)
  medoid_idx <- which.min(rowMeans(m))
  sample_ids[medoid_idx]
}

# Identify all T0 samples and compute global T0 medoid
t0_meta <- meta %>% filter(Time_Point == "T0")
t0_ids  <- t0_meta$SampleID
global_medoid <- get_medoid_id(t0_ids)

if (is.na(global_medoid)) {
  stop("No T0 samples found to define a global baseline.")
}

# ---- Compute Bray–Curtis vs GLOBAL T0 baseline (T0 medoid = 0) ----
compute_dist_to_globalT0 <- function(sid, global_id) {
  if (is.na(sid) || is.na(global_id)) return(NA_real_)
  if (sid == global_id) return(0)  # global T0 medoid: distance = 0
  as.numeric(
    vegdist(
      rbind(rel[global_id, , drop = FALSE],
            rel[sid,        , drop = FALSE]),
      method = "bray"
    )
  )[1]
}

dist_df <- meta %>%
  mutate(
    Bray_to_GlobalT0 = map_dbl(SampleID, ~ compute_dist_to_globalT0(.x, global_medoid)),
    Time_Point       = factor(Time_Point, levels = time_points),
    TreatmentGroup   = factor(TreatmentGroup, levels = treatments)
  )

# ---- Numeric x for LOESS ----
dist_df <- dist_df %>%
  mutate(tp_num = as.integer(factor(Time_Point, levels = c("T0","T2","T3","T6"))))

# ---- Plot ----
p <- ggplot(dist_df, aes(x = tp_num, y = Bray_to_GlobalT0, color = Time_Point)) +
  geom_point(
    position = position_jitter(width = 0.08, height = 0),
    alpha = 0.75, size = 2.5
  ) +
  geom_smooth(
    aes(group = 1),
    method = "loess", se = TRUE, span = 0.9,
    linewidth = 1.0, color = "black"
  ) +
  scale_x_continuous(
    breaks = 1:4,
    labels = c("T0","T2","T3","T6"),
    limits = c(1,4)
  ) +
  scale_color_manual(values = time_colors, drop = FALSE) +
  facet_wrap(~ TreatmentGroup, ncol = 3, drop = FALSE) +
  labs(
    title = "Bray–Curtis Dissimilarity vs Global T0 Baseline",
    x = "Time Point",
    y = "Bray–Curtis vs Global T0 Medoid"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(face = "bold", size = 16),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )

print(p)

# ---- Save ----
ggsave(
  file.path(root.dir, "BrayCurtis_vs_GlobalT0_Dotplot_byTreatment.pdf"),
  p, width = 14, height = 9, device = "pdf"
)
ggsave(
  file.path(root.dir, "BrayCurtis_vs_GlobalT0_Dotplot_byTreatment.tiff"),
  p, width = 14, height = 9, dpi = 300, device = "tiff"
)

```

PLSDA of BetaDiversity
```{r pcoa_plots, fig.width=15, fig.height=10}
library(phyloseq)
library(ggplot2)
library(patchwork)
library(ggforce)
library(mixOmics)
library(vegan)  # for decostand (e.g., Hellinger)
library(tibble)
library(dplyr)

# Filter time points you want
time_points <- c("T0", "T2", "T3", "T6")  # changed here

plsda_plots <- vector("list", length(time_points))

for (i in seq_along(time_points)) {
  tp <- time_points[i]
  
  # Subset ps.rarefied to current time point
  ps_sub <- subset_samples(ps.rarefied, Time_Point == tp)
  ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)  # Drop zero taxa
  
  # Extract ASV count data
  asv <- as.data.frame(otu_table(ps_sub))
  if (taxa_are_rows(ps_sub)) {
    asv <- t(asv)
  }
  
  # Hellinger transform (recommended for ecological data)
  asv_hel <- decostand(asv, method = "hellinger")
  
  metadata <- as.data.frame(sample_data(ps_sub))
  group <- metadata$TreatmentGroup
  
  # Run PLS-DA
  plsda_model <- plsda(asv_hel, group, ncomp = 2)
  
  # Extract component scores
  scores <- as.data.frame(plsda_model$variates$X)
  scores$TreatmentGroup <- group
  scores$SampleID <- rownames(scores)
  
  # Axis limits with 10% buffer
  x_range <- range(scores$comp1)
  y_range <- range(scores$comp2)
  x_buffer <- diff(x_range) * 0.1
  y_buffer <- diff(y_range) * 0.1
  x_limits <- c(x_range[1] - x_buffer, x_range[2] + x_buffer)
  y_limits <- c(y_range[1] - y_buffer, y_range[2] + y_buffer)
  
  # Plot for this time point
  plsda_plots[[i]] <- ggplot(scores, aes(x = comp1, y = comp2, color = TreatmentGroup)) +
    geom_point(size = 3, alpha = 0.7) +
    ggforce::geom_mark_ellipse(aes(group = TreatmentGroup, fill = TreatmentGroup), alpha = 0.1, show.legend = FALSE) +
    labs(title = paste("PLS-DA -", tp), x = "Component 1", y = "Component 2") +
    coord_cartesian(xlim = x_limits, ylim = y_limits) +
    theme_minimal() +
    theme(
      legend.position = ifelse(i == length(time_points), "right", "none"),
      plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm")
    )
}

# Combine plots with shared legend
combined_plsda <- wrap_plots(plsda_plots, ncol = 2, nrow = 2) + 
  plot_layout(guides = "collect") &
  theme(plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm"))

print(combined_plsda)
```

# Save to file in root.dir
ggsave(
  filename = file.path(root.dir, "PLSDA_Filtered_TimePoints.pdf"),
  plot = combined_plsda,
  width = 15,
  height = 10,
  device = "pdf"
)

ggsave(
  filename = file.path(root.dir, "PLSDA_Filtered_TimePoints.tiff"),
  plot = combined_plsda,
  width = 15,
  height = 10,
  dpi = 300,
  device = "tiff"
)


```

```{r}

# Convert taxonomy to data frame
tax_df <- as.data.frame(tax_table(ps))

# Separate concatenated taxonomy string into levels
library(tidyr)
tax_df <- tax_df %>%
  separate(Taxon, into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";", fill = "right")

# Replace tax_table in the phyloseq object
tax_table(ps) <- tax_table(as.matrix(tax_df))
rank_names(ps)

```


4. Differential Abundance Analysis
```{r}
library(phyloseq)
library(DESeq2)
library(tidyverse)
library(ggrepel)
library(knitr)

# Ensure proper factor levels
sample_data(ps)$TreatmentGroup <- factor(sample_data(ps)$TreatmentGroup)
sample_data(ps)$Time_Point <- factor(sample_data(ps)$Time_Point)

# Extract taxonomy table
tax_df <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("FeatureID")

# Output directory
deseq_dir <- file.path(root.dir, "DESeq2_Results")
if (!dir.exists(deseq_dir)) dir.create(deseq_dir)

# Filter time points to only desired ones
desired_time_points <- c("T0", "T2", "T3", "T6")
time_points <- intersect(levels(sample_data(ps)$Time_Point), desired_time_points)
treatments <- setdiff(levels(sample_data(ps)$TreatmentGroup), "Vehicle")

summary_list <- list()

for (tp in time_points) {
  for (treat in treatments) {
    
    ps_sub <- subset_samples(ps, Time_Point == tp)
    ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)

    groups <- unique(sample_data(ps_sub)$TreatmentGroup)
    if (!all(c("Vehicle", treat) %in% groups) || nsamples(ps_sub) < 4) next
    
    sample_data(ps_sub)$TreatmentGroup <- relevel(factor(sample_data(ps_sub)$TreatmentGroup), ref = "Vehicle")

    dds <- phyloseq_to_deseq2(ps_sub, ~ TreatmentGroup)
    keep <- rowSums(counts(dds)) > 0
    dds <- dds[keep, ]
    dds <- DESeq(dds, sfType = "poscounts")

    res <- results(dds, contrast = c("TreatmentGroup", treat, "Vehicle"))
    
    # Join taxonomy and calculate summary columns
    res_df <- as.data.frame(res) %>%
      rownames_to_column("FeatureID") %>%
      left_join(tax_df, by = "FeatureID")

    taxonomy_levels <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
    available_levels <- intersect(taxonomy_levels, colnames(res_df))

    res_df <- res_df %>%
      mutate(
        neglog10padj = -log10(padj),
        regulation = case_when(
          padj < 0.05 & log2FoldChange > 1  ~ "Up",
          padj < 0.05 & log2FoldChange < -1 ~ "Down",
          TRUE                              ~ "NS"
        ),
        label = ifelse(
          padj < 0.05 & abs(log2FoldChange) > 1,
          apply(dplyr::select(., all_of(available_levels)), 1, function(x) {
            x <- as.character(x)
            x <- trimws(tolower(x))
            x <- x[!x %in% c("none", "unclassified", "unknown", "", NA)]
            if (length(x) == 0) NA else tail(x, 1)
          }),
          ""
        )
      )

    sig_df <- res_df %>% filter(!is.na(padj) & padj < 0.05)

    comp_dir <- file.path(deseq_dir, paste0("Time_", tp, "_", treat, "_vs_Vehicle"))
    if (!dir.exists(comp_dir)) dir.create(comp_dir, recursive = TRUE)

    # Save results
    write.csv(res_df, file = file.path(comp_dir, "all_results.csv"), row.names = FALSE)
    write.csv(sig_df, file = file.path(comp_dir, "significant_results.csv"), row.names = FALSE)
    saveRDS(dds, file = file.path(comp_dir, "dds_object.rds"))

    # Base volcano plot
    volcano_base <- ggplot(res_df, aes(x = log2FoldChange, y = neglog10padj)) +
      geom_point(aes(color = regulation), alpha = 0.7, size = 1.5) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
      geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "gray50") +
      scale_color_manual(values = c("Up" = "firebrick", "Down" = "dodgerblue", "NS" = "gray70")) +
      labs(
        title = paste("Volcano Plot:", treat, "vs Vehicle at", tp),
        x = "log2 Fold Change",
        y = "-log10 Adjusted p-value"
      ) +
      theme_minimal() +
      theme(legend.position = "none")

    # Top 10 upregulated by padj
    top_up <- res_df %>%
      filter(!is.na(padj), padj < 0.05, log2FoldChange > 0) %>%
      arrange(padj) %>%
      slice_head(n = 10)

    # Top 10 downregulated by padj
    top_down <- res_df %>%
      filter(!is.na(padj), padj < 0.05, log2FoldChange < 0) %>%
      arrange(padj) %>%
      slice_head(n = 10)

    top10 <- bind_rows(top_up, top_down)

    # Annotated volcano plot
    volcano_annotated <- volcano_base +
      ggrepel::geom_text_repel(
        data = top10,
        aes(label = label),
        size = 3,
        max.overlaps = 20,
        box.padding = 0.3,
        segment.color = "gray50"
      )

    # Save volcano plots inside comp_dir
    ggsave(file.path(comp_dir, "volcano_plot_annotated.pdf"), plot = volcano_annotated, width = 6, height = 5)
    ggsave(file.path(comp_dir, "volcano_plot_annotated.tiff"), plot = volcano_annotated, width = 6, height = 5, dpi = 300)

    ggsave(file.path(comp_dir, "volcano_plot_plain.pdf"), plot = volcano_base, width = 6, height = 5)
    ggsave(file.path(comp_dir, "volcano_plot_plain.tiff"), plot = volcano_base, width = 6, height = 5, dpi = 300)

    # Add to summary
    top_taxon_cols <- intersect(c("Genus", "Family", "Phylum"), colnames(sig_df))
    top_taxon <- if (nrow(sig_df) > 0 && length(top_taxon_cols) > 0) {
      paste(na.omit(sig_df[order(sig_df$padj), ][1, top_taxon_cols]), collapse = "; ")
    } else {
      NA
    }

    summary_list[[length(summary_list) + 1]] <- tibble(
      Time_Point = tp,
      Treatment = treat,
      Num_Significant = nrow(sig_df),
      Top_Taxon = top_taxon
    )
  }
}

# Save and display summary table
summary_table <- bind_rows(summary_list)
write.csv(summary_table, file = file.path(deseq_dir, "DESeq2_summary_table.csv"), row.names = FALSE)
knitr::kable(summary_table, caption = "Summary of DESeq2 Results by Treatment and Time Point")

```

```{r summary_pdf_of_volcano_plots, message=FALSE, warning=FALSE}
library(pdftools)
library(grid)
library(gridExtra)
library(magick)
library(purrr)

# Set directory with DESeq2 volcano plots
deseq_dir <- file.path(root.dir, "DESeq2_Results")

# Find all annotated volcano PDF files recursively
volcano_pdf_files <- list.files(deseq_dir, pattern = "volcano_plot_annotated\\.pdf$", recursive = TRUE, full.names = TRUE)

# Read and convert first page of each PDF to magick image
volcano_images <- map(volcano_pdf_files, ~ image_read_pdf(.x, pages = 1))

# Resize for consistency
volcano_images_resized <- map(volcano_images, ~ image_scale(.x, "600"))

# Convert magick images to grobs for grid layout
volcano_grobs <- map(volcano_images_resized, ~ rasterGrob(as.raster(.x)))

# Define layout per page
plots_per_page <- 6  # 3 columns × 2 rows
ncol <- 3
nrow <- 2
n_pages <- ceiling(length(volcano_grobs) / plots_per_page)

# Save paginated summary PDF
summary_pdf_path <- file.path(root.dir, "Volcano_Plots_Summary.pdf")
pdf(summary_pdf_path, width = 18, height = 12)  # Wider than tall
for (i in seq_len(n_pages)) {
  start_idx <- (i - 1) * plots_per_page + 1
  end_idx <- min(i * plots_per_page, length(volcano_grobs))
  grid.arrange(grobs = volcano_grobs[start_idx:end_idx], ncol = ncol)
}
dev.off()

cat("✅ Volcano summary PDF saved to:", summary_pdf_path, "\n")



```


# Differential Abundance and Volcano Plots comparing timepoints within each treatment group
```{r}
# Required libraries
library(phyloseq)
library(DESeq2)
library(tidyverse)
library(ggrepel)
library(knitr)
library(patchwork)

# Set factors
sample_data(ps)$TreatmentGroup <- factor(sample_data(ps)$TreatmentGroup)
sample_data(ps)$Time_Point <- factor(sample_data(ps)$Time_Point)

# Taxonomy
tax_df <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("FeatureID")

# Output root
timecourse_dir <- file.path(root.dir, "DESeq2_Results_Timecourse_AllPairs")
if (!dir.exists(timecourse_dir)) dir.create(timecourse_dir)

# Time points and treatment groups
all_times <- levels(sample_data(ps)$Time_Point)
treatments <- levels(sample_data(ps)$TreatmentGroup)
time_pairs <- combn(all_times, 2, simplify = FALSE)

summary_list <- list()
volcano_plots <- list()

for (treat in treatments) {
  treat_dir <- file.path(timecourse_dir, paste0("Treatment_", treat))
  if (!dir.exists(treat_dir)) dir.create(treat_dir)

  for (tp_pair in time_pairs) {
    tp1 <- tp_pair[1]
    tp2 <- tp_pair[2]

    cat("Processing", treat, tp2, "vs", tp1, "\n")

    # Subset
    ps_sub <- subset_samples(ps, TreatmentGroup == treat & Time_Point %in% c(tp1, tp2))
    ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)

    if (nsamples(ps_sub) < 4 || length(unique(sample_data(ps_sub)$Time_Point)) < 2) next

    sample_data(ps_sub)$Time_Point <- relevel(factor(sample_data(ps_sub)$Time_Point), ref = tp1)

    dds <- phyloseq_to_deseq2(ps_sub, ~ Time_Point)
    keep <- rowSums(counts(dds)) > 0
    dds <- dds[keep, ]
    dds <- DESeq(dds, sfType = "poscounts")

    res <- results(dds, contrast = c("Time_Point", tp2, tp1))

    res_df <- as.data.frame(res) %>%
      rownames_to_column("FeatureID") %>%
      left_join(tax_df, by = "FeatureID")

    taxonomy_levels <- c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")
    available_levels <- intersect(taxonomy_levels, colnames(res_df))

    res_df <- res_df %>%
      mutate(
        neglog10padj = -log10(padj),
        regulation = case_when(
          padj < 0.05 & log2FoldChange > 1  ~ "Up",
          padj < 0.05 & log2FoldChange < -1 ~ "Down",
          TRUE                              ~ "NS"
        ),
        label = ifelse(
          padj < 0.05 & abs(log2FoldChange) > 1,
          apply(dplyr::select(., all_of(available_levels)), 1, function(x) {
            x <- as.character(x)
            x <- trimws(tolower(x))
            x <- x[!x %in% c("none", "unclassified", "unknown", "", NA)]
            if (length(x) == 0) NA else tail(x, 1)
          }),
          ""
        )
      )

    sig_df <- res_df %>% filter(!is.na(padj) & padj < 0.05)

    comp_dir <- file.path(treat_dir, paste0(tp2, "_vs_", tp1))
    if (!dir.exists(comp_dir)) dir.create(comp_dir, recursive = TRUE)

    # Save data
    write.csv(res_df, file = file.path(comp_dir, "all_results.csv"), row.names = FALSE)
    write.csv(sig_df, file = file.path(comp_dir, "significant_results.csv"), row.names = FALSE)
    saveRDS(dds, file = file.path(comp_dir, "dds_object.rds"))

    # Volcano plot
    volcano_base <- ggplot(res_df, aes(x = log2FoldChange, y = neglog10padj)) +
      geom_point(aes(color = regulation), alpha = 0.7, size = 1.5) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
      geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "gray50") +
      scale_color_manual(values = c("Up" = "firebrick", "Down" = "dodgerblue", "NS" = "gray70")) +
      labs(
        title = paste("Volcano:", treat, tp2, "vs", tp1),
        x = "log2 Fold Change",
        y = "-log10 Adjusted p-value"
      ) +
      theme_minimal() +
      theme(legend.position = "none")

    top10 <- res_df %>%
      filter(!is.na(padj), padj < 0.05) %>%
      arrange(padj) %>%
      slice_head(n = 10)

    volcano_annotated <- volcano_base +
      ggrepel::geom_text_repel(
        data = top10,
        aes(label = label),
        size = 3,
        max.overlaps = 10,
        box.padding = 0.3
      )

    # Save volcano plots
    ggsave(file.path(comp_dir, "volcano_plot_annotated.pdf"), plot = volcano_annotated, width = 6, height = 5)
    ggsave(file.path(comp_dir, "volcano_plot_plain.pdf"), plot = volcano_base, width = 6, height = 5)
    ggsave(file.path(comp_dir, "volcano_plot_annotated.tiff"), plot = volcano_annotated, width = 6, height = 5, dpi = 300)
    ggsave(file.path(comp_dir, "volcano_plot_plain.tiff"), plot = volcano_base, width = 6, height = 5, dpi = 300)

    # Summary info
    num_significant <- nrow(sig_df)
    fdr_stat <- sum(sig_df$padj < 0.05, na.rm = TRUE)

    top_taxon_cols <- intersect(c("Genus", "Family", "Phylum"), colnames(sig_df))
    top_taxon <- if (num_significant > 0 && length(top_taxon_cols) > 0) {
      paste(na.omit(sig_df[order(sig_df$padj), ][1, top_taxon_cols]), collapse = "; ")
    } else {
      NA
    }

    summary_list[[length(summary_list) + 1]] <- tibble(
      Treatment = treat,
      Comparison = paste(tp2, "vs", tp1),
      Num_Significant = num_significant,
      Top_Taxon = top_taxon,
      FDR_Less_0_05 = fdr_stat
    )

    # Collect volcano for summary
    volcano_plots[[length(volcano_plots) + 1]] <- volcano_annotated
  }
}

# Multi-page volcano summary PDF
summary_pdf <- file.path(timecourse_dir, "volcano_summary_all_pairs.pdf")
n_per_page <- 6
n_pages <- ceiling(length(volcano_plots) / n_per_page)

pdf(summary_pdf, width = 14, height = 10)
for (i in seq_len(n_pages)) {
  idx_start <- (i - 1) * n_per_page + 1
  idx_end <- min(i * n_per_page, length(volcano_plots))
  plots_page <- volcano_plots[idx_start:idx_end]
  patchwork::wrap_plots(plots_page, ncol = 3, nrow = 2) %>% print()
}
dev.off()

# Save summary table
summary_table <- bind_rows(summary_list)
write.csv(summary_table, file = file.path(timecourse_dir, "DESeq2_timecourse_summary_table.csv"), row.names = FALSE)
knitr::kable(summary_table, caption = "Summary of DESeq2 Pairwise Time Point Comparisons Within Each Treatment Group")

```


## Differential abundance at multiple taxonomic ranks:
## Timepoint comparisons within each treatment group, volcano plots
## Each point = one taxon at given rank (Class, Order, Family, Genus, Species)

```{r}
library(phyloseq)
library(DESeq2)
library(tidyverse)
library(ggrepel)

# Make sure factors are set
sample_data(ps)$TreatmentGroup <- factor(sample_data(ps)$TreatmentGroup)
sample_data(ps)$Time_Point     <- factor(sample_data(ps)$Time_Point)

# Taxonomic ranks to analyze
rank_vec <- c("Class", "Order", "Family", "Genus", "Species")

# Base output root for rank-level results
timecourse_rank_root <- file.path(root.dir, "DESeq2_Results_Timecourse_ByRank")
if (!dir.exists(timecourse_rank_root)) dir.create(timecourse_rank_root, recursive = TRUE)

# Time points and treatments (reuse your levels)
all_times  <- levels(sample_data(ps)$Time_Point)
treatments <- levels(sample_data(ps)$TreatmentGroup)
time_pairs <- combn(all_times, 2, simplify = FALSE)

for (rank_name in rank_vec) {

  message("#############################")
  message("### Rank: ", rank_name)
  message("#############################")

  # Check that this rank exists in the taxonomy
  tax_cols <- colnames(tax_table(ps))
  if (!(rank_name %in% tax_cols)) {
    message("  - Rank ", rank_name, " not found in tax_table(ps); skipping.")
    next
  }

  # Rank-specific output directory
  rank_dir_root <- file.path(timecourse_rank_root, rank_name)
  if (!dir.exists(rank_dir_root)) dir.create(rank_dir_root, recursive = TRUE)

  for (treat in treatments) {
    message("====== Treatment: ", treat, " ======")

    for (tp_pair in time_pairs) {
      tp1 <- tp_pair[1]
      tp2 <- tp_pair[2]

      message("  Comparing ", tp2, " vs ", tp1)

      # Subset to this treatment and these two timepoints
      ps_sub <- subset_samples(ps, TreatmentGroup == treat & Time_Point %in% c(tp1, tp2))
      ps_sub <- prune_taxa(taxa_sums(ps_sub) > 0, ps_sub)

      if (nsamples(ps_sub) < 4 || length(unique(sample_data(ps_sub)$Time_Point)) < 2) {
        message("    - Not enough samples or only one timepoint present, skipping.")
        next
      }

      # ---- Aggregate to current taxonomic rank ----
      ps_rank <- tax_glom(ps_sub, taxrank = rank_name)
      ps_rank <- prune_taxa(taxa_sums(ps_rank) > 0, ps_rank)

      # Relevel timepoint so tp1 is reference
      sample_data(ps_rank)$Time_Point <- relevel(
        factor(sample_data(ps_rank)$Time_Point),
        ref = tp1
      )

      # DESeq2 at this rank
      dds <- phyloseq_to_deseq2(ps_rank, ~ Time_Point)
      keep <- rowSums(counts(dds)) > 0
      dds  <- dds[keep, ]

      if (nrow(dds) == 0) {
        message("    - No taxa left after filtering for ", rank_name, "; skipping.")
        next
      }

      dds  <- DESeq(dds, sfType = "poscounts")
      res  <- results(dds, contrast = c("Time_Point", tp2, tp1))

      # Taxonomy table at this rank
      tax_df <- as.data.frame(tax_table(ps_rank)) %>%
        rownames_to_column("FeatureID")

      # Build results table
      res_df <- as.data.frame(res) %>%
        rownames_to_column("FeatureID") %>%
        left_join(tax_df, by = "FeatureID") %>%
        mutate(
          neglog10padj = -log10(padj),
          regulation = case_when(
            padj < 0.05 & log2FoldChange >  1 ~ "Up",
            padj < 0.05 & log2FoldChange < -1 ~ "Down",
            TRUE                              ~ "NS"
          ),
          # label with the name at this rank
          label = !!rlang::sym(rank_name)
        )

      # Make comparison-specific directory
      comp_dir <- file.path(
        rank_dir_root,
        paste0("Treatment_", treat, "_", tp2, "_vs_", tp1)
      )
      if (!dir.exists(comp_dir)) dir.create(comp_dir, recursive = TRUE)

      # Save table + dds
      write.csv(res_df,
                file = file.path(comp_dir, paste0(tolower(rank_name), "_level_all_results.csv")),
                row.names = FALSE)
      saveRDS(dds,
              file = file.path(comp_dir, paste0(tolower(rank_name), "_level_dds_object.rds")))

      # Volcano: each point = one taxon at this rank
      volcano_base <- ggplot(res_df, aes(x = log2FoldChange, y = neglog10padj)) +
        geom_point(aes(color = regulation), alpha = 0.8, size = 2) +
        geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
        geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "gray50") +
        scale_color_manual(values = c("Up" = "firebrick", "Down" = "dodgerblue", "NS" = "gray70")) +
        labs(
          title = paste(rank_name, "level DA:", treat, "-", tp2, "vs", tp1),
          x = paste("log2 Fold Change (", rank_name, ")", sep = ""),
          y = "-log10 Adjusted p-value"
        ) +
        theme_minimal() +
        theme(legend.position = "none")

      # Label top 10 taxa at this rank by smallest padj
      top10 <- res_df %>%
        filter(!is.na(padj)) %>%
        arrange(padj) %>%
        slice_head(n = 10)

      volcano_annotated <- volcano_base +
        ggrepel::geom_text_repel(
          data = top10,
          aes(label = label),
          size = 3,
          max.overlaps = 10,
          box.padding = 0.3
        )

      # Save plots (PDF + PNG, avoid TIFF issues)
      ggsave(file.path(comp_dir, paste0("volcano_", tolower(rank_name), "_level_annotated.pdf")),
             plot = volcano_annotated, width = 6, height = 5)
      ggsave(file.path(comp_dir, paste0("volcano_", tolower(rank_name), "_level_plain.pdf")),
             plot = volcano_base, width = 6, height = 5)

      ggsave(file.path(comp_dir, paste0("volcano_", tolower(rank_name), "_level_annotated.png")),
             plot = volcano_annotated, width = 6, height = 5, dpi = 300)
      ggsave(file.path(comp_dir, paste0("volcano_", tolower(rank_name), "_level_plain.png")),
             plot = volcano_base, width = 6, height = 5, dpi = 300)
    }
  }
}

message("✅ Done! Rank-level timecourse volcano plots saved in:\n",
        timecourse_rank_root, "\n")
```

This chunk creates volcano plots of each individual Class and order - no needed
```{r}
## Volcano plots by Class and Order using existing DESeq2 results
## Place this chunk AFTER your DESeq2 code that creates all_results.csv

library(tidyverse)
library(ggrepel)
library(grDevices)
options(bitmapType = "cairo")

# ---- Input DESeq2 result directories ----
input_dirs <- c(
  file.path(root.dir, "DESeq2_Results"),
  file.path(root.dir, "DESeq2_Results_Timecourse_AllPairs")
)
input_dirs <- input_dirs[dir.exists(input_dirs)]

if (length(input_dirs) == 0) {
  stop("No DESeq2 result directories found under root.dir. Check root.dir or run DESeq2 chunks first.")
}

# ---- Output roots for Class/Order volcanoes ----
class_root <- file.path(root.dir, "DESeq2_Volcano_By_Class")
order_root <- file.path(root.dir, "DESeq2_Volcano_By_Order")
if (!dir.exists(class_root)) dir.create(class_root, recursive = TRUE)
if (!dir.exists(order_root)) dir.create(order_root, recursive = TRUE)

# ---- Global summary tables across ALL comparisons ----
class_summary_global <- tibble()
order_summary_global <- tibble()

# ---- Helper: make volcano plots restricted to one taxonomic level (Class or Order) ----
make_tax_volcanos <- function(res_df, comparison_id, tax_level, out_root,
                              min_features_for_plot = 5) {
  message(">>> ", comparison_id, " | tax_level = ", tax_level)

  if (!(tax_level %in% colnames(res_df))) {
    message("  - Column ", tax_level, " not found in res_df; skipping.")
    return(invisible(NULL))
  }

  # Folder for this comparison under the chosen root
  comp_root <- file.path(out_root, comparison_id)
  if (!dir.exists(comp_root)) dir.create(comp_root, recursive = TRUE)

  # Get unique taxa at this level
  taxa_vals <- sort(unique(res_df[[tax_level]]))
  taxa_vals <- taxa_vals[!is.na(taxa_vals) & taxa_vals != ""]

  message("  - Found ", length(taxa_vals), " taxa at level ", tax_level)

  comparison_summary <- tibble()

  for (tv in taxa_vals) {
    sub_df <- res_df[res_df[[tax_level]] == tv & !is.na(res_df$padj), ]
    n_feat <- nrow(sub_df)
    message("    * Taxon: ", tv, " | n_features = ", n_feat)

    if (n_feat == 0) next

    # Ensure needed columns exist
    if (!"neglog10padj" %in% colnames(sub_df)) {
      sub_df$neglog10padj <- -log10(sub_df$padj)
    }
    if (!"regulation" %in% colnames(sub_df)) {
      sub_df <- sub_df %>%
        mutate(
          regulation = case_when(
            padj < 0.05 & log2FoldChange >  1 ~ "Up",
            padj < 0.05 & log2FoldChange < -1 ~ "Down",
            TRUE                              ~ "NS"
          )
        )
    }
    if (!"label" %in% colnames(sub_df)) {
      # If no label column, create something from taxonomy
      tax_cols <- intersect(c("Species", "Genus", "Family", "Order", "Class", "Phylum"), colnames(sub_df))
      if (length(tax_cols) > 0) {
        sub_df <- sub_df %>%
          mutate(
            label = apply(dplyr::select(., all_of(tax_cols)), 1, function(x) {
              x <- as.character(x)
              x <- trimws(x)
              x <- x[!is.na(x) & x != "" & !tolower(x) %in% c("none", "unclassified", "unknown")]
              if (length(x) == 0) "" else tail(x, 1)
            })
          )
      } else {
        sub_df$label <- ""
      }
    }

    # ---- summary row (always recorded, even if we don't plot) ----
    summary_row <- tibble(
      comparison_id   = comparison_id,
      tax_level       = tax_level,
      taxon           = tv,
      Num_Features    = n_feat,
      Num_Significant = sum(sub_df$padj < 0.05, na.rm = TRUE),
      Num_Up          = sum(sub_df$padj < 0.05 & sub_df$log2FoldChange >  1, na.rm = TRUE),
      Num_Down        = sum(sub_df$padj < 0.05 & sub_df$log2FoldChange < -1, na.rm = TRUE)
    )

    comparison_summary <- dplyr::bind_rows(comparison_summary, summary_row)

    if (tax_level == "Class") {
      class_summary_global <<- dplyr::bind_rows(class_summary_global, summary_row)
    } else if (tax_level == "Order") {
      order_summary_global <<- dplyr::bind_rows(order_summary_global, summary_row)
    }

    # ---- plotting only if enough features ----
    if (n_feat < min_features_for_plot) {
      message("      - Less than ", min_features_for_plot, " features; summarizing but skipping plots.")
      next
    }

    safe_name <- gsub("[^A-Za-z0-9_]+", "_", tv)
    tax_dir <- file.path(comp_root, paste0(tax_level, "_", safe_name))
    if (!dir.exists(tax_dir)) dir.create(tax_dir, recursive = TRUE)

    volcano_base <- ggplot(sub_df, aes(x = log2FoldChange, y = neglog10padj)) +
      geom_point(aes(color = regulation), alpha = 0.7, size = 1.5) +
      geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
      geom_vline(xintercept = c(-1, 1), linetype = "dotted", color = "gray50") +
      scale_color_manual(values = c("Up" = "firebrick", "Down" = "dodgerblue", "NS" = "gray70")) +
      labs(
        title = paste0("Volcano: ", comparison_id, " | ", tax_level, " = ", tv),
        x = "log2 Fold Change",
        y = "-log10 Adjusted p-value"
      ) +
      theme_minimal() +
      theme(legend.position = "none")

    top10 <- sub_df %>%
      dplyr::filter(!is.na(padj), padj < 0.05) %>%
      dplyr::arrange(padj) %>%
      dplyr::slice_head(n = 10)

    volcano_annotated <- volcano_base +
      ggrepel::geom_text_repel(
        data = top10,
        aes(label = label),
        size = 3,
        max.overlaps = 10,
        box.padding = 0.3
      )

    message("      - Saving plots to: ", tax_dir)
    write.csv(sub_df,
              file = file.path(tax_dir, "all_results_taxon_subset.csv"),
              row.names = FALSE)

    ggsave(file.path(tax_dir, "volcano_plot_annotated.pdf"),
           plot = volcano_annotated, width = 6, height = 5)
    ggsave(file.path(tax_dir, "volcano_plot_annotated.tiff"),
           plot = volcano_annotated, width = 6, height = 5, dpi = 300)
    ggsave(file.path(tax_dir, "volcano_plot_plain.pdf"),
           plot = volcano_base, width = 6, height = 5)
    ggsave(file.path(tax_dir, "volcano_plot_plain.tiff"),
           plot = volcano_base, width = 6, height = 5, dpi = 300)
  }

  # Per-comparison summary CSV
  if (nrow(comparison_summary) > 0) {
    write.csv(comparison_summary,
              file = file.path(comp_root, paste0("summary_by_", tax_level, ".csv")),
              row.names = FALSE)
  }

  invisible(NULL)
}

# ---- Find all all_results.csv files from your DESeq2 runs ----
all_result_files <- unlist(lapply(
  input_dirs,
  function(d) list.files(d, pattern = "all_results\\.csv$", recursive = TRUE, full.names = TRUE)
))

if (length(all_result_files) == 0) {
  stop("No all_results.csv files found under DESeq2_Results directories.")
}

# ---- Loop over each all_results.csv, make Class/Order volcanoes ----
for (f in all_result_files) {
  message("=== Processing file: ", f)

  res_df <- read.csv(f, check.names = FALSE)

  # Need at least these columns to proceed
  if (!all(c("log2FoldChange", "padj") %in% colnames(res_df))) {
    message("  - Missing log2FoldChange or padj in file; skipping.")
    next
  }

  # Ensure neglog10padj exists
  if (!"neglog10padj" %in% colnames(res_df)) {
    res_df$neglog10padj <- -log10(res_df$padj)
  }

  # Ensure regulation exists
  if (!"regulation" %in% colnames(res_df)) {
    res_df <- res_df %>%
      mutate(
        regulation = case_when(
          padj < 0.05 & log2FoldChange >  1 ~ "Up",
          padj < 0.05 & log2FoldChange < -1 ~ "Down",
          TRUE                              ~ "NS"
        )
      )
  }

  # Define a comparison_id from folder structure
  comp_dir  <- dirname(f)
  comp_name <- basename(comp_dir)                 # e.g. Time_T2_Quetiapine_vs_Vehicle
  parent    <- basename(dirname(comp_dir))        # e.g. Time_T2_Quetiapine_vs_Vehicle's parent
  comparison_id <- paste(parent, comp_name, sep = "__")

  # Create plots & summaries by Class and Order
  make_tax_volcanos(res_df, comparison_id, "Class", class_root)
  make_tax_volcanos(res_df, comparison_id, "Order", order_root)
}

# ---- Write global summary tables ----
if (nrow(class_summary_global) > 0) {
  write.csv(class_summary_global,
            file = file.path(class_root, "Class_DESeq2_summary_across_comparisons.csv"),
            row.names = FALSE)
}

if (nrow(order_summary_global) > 0) {
  write.csv(order_summary_global,
            file = file.path(order_root, "Order_DESeq2_summary_across_comparisons.csv"),
            row.names = FALSE)
}

message("✅ Done! Class/Order volcano plots and summaries saved in:\n",
        "  - ", class_root, "\n",
        "  - ", order_root, "\n")


```





Define Color Palette and make bar chart
```{r}
library(phyloseq)
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# Filter and prune samples
included_timepoints <- c("T0", "T2", "T3", "T6")
ps_filtered <- subset_samples(ps, Time_Point %in% included_timepoints)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# Agglomerate to Phylum level
ps_phylum <- tax_glom(ps_filtered, taxrank = "Phylum")

# Extract and clean taxonomy matrix
tax_mat <- as.matrix(tax_table(ps_phylum))
tax_mat[, "Phylum"] <- trimws(tax_mat[, "Phylum"])
tax_mat[, "Phylum"][tax_mat[, "Phylum"] == "none"] <- "Other"

# IMPORTANT: store Phylum as CHARACTER (not factor) to keep labels in tax_table
tax_mat[, "Phylum"] <- as.character(tax_mat[, "Phylum"])

# Assign cleaned taxonomy back
tax_table(ps_phylum) <- tax_table(tax_mat)

# Define palette
phyla_all <- sort(unique(tax_mat[, "Phylum"]))
n_colors <- max(3, min(12, length(phyla_all)))
palette_colors <- brewer.pal(n_colors, "Set3")
phylum_colors <- setNames(palette_colors[1:length(phyla_all)], phyla_all)

# Melt for plotting
df_bar <- psmelt(ps_phylum)

# Check that Phylum is character with proper labels (no numeric codes)
print(unique(df_bar$Phylum))

# Plot with manual fill
p_bar <- ggplot(df_bar, aes(x = Time_Point, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~ TreatmentGroup) +
  theme_minimal() +
  labs(title = "Taxonomic Composition at Phylum Level (T0, T2, T3, T6)") +
  scale_fill_manual(values = phylum_colors) +
  theme(
    plot.title = element_text(size = 20, face = "bold", margin = margin(b = 15)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

print(p_bar)

# Save plot
ggsave(file.path(root.dir, "phylum_barplot_fixed.pdf"), plot = p_bar, width = 10, height = 6)
ggsave(file.path(root.dir, "phylum_barplot_fixed.tiff"), plot = p_bar, width = 10, height = 6, dpi = 300)



```
```{r}
library(phyloseq)
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# 1️⃣ Filter and prune samples
included_timepoints <- c("T0", "T2", "T3", "T6")
ps_filtered <- subset_samples(ps, Time_Point %in% included_timepoints)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# 2️⃣ Agglomerate to Phylum
ps_phylum <- tax_glom(ps_filtered, taxrank = "Phylum")

# 3️⃣ Clean taxonomy table
tax_mat <- as.matrix(tax_table(ps_phylum))
tax_mat[, "Phylum"] <- trimws(tax_mat[, "Phylum"])
tax_mat[, "Phylum"][tax_mat[, "Phylum"] == "none"] <- "Other"
tax_mat[, "Phylum"] <- as.character(tax_mat[, "Phylum"])
tax_table(ps_phylum) <- tax_table(tax_mat)

# 4️⃣ Transform to relative abundance before melting
ps_rel <- transform_sample_counts(ps_phylum, function(x) x / sum(x))

# 5️⃣ Melt into long format
df_bar <- psmelt(ps_rel)

# Convert to percentage
df_bar$RelAbundance <- df_bar$Abundance * 100

# 🔍 Verify that each sample sums to 100%
check_sum <- df_bar %>%
  group_by(Sample) %>%
  summarise(total = sum(RelAbundance))
print(summary(check_sum$total))  # all should be ≈100

# 6️⃣ Optionally average within TreatmentGroup × Time_Point
# (useful if you have replicates and want one bar per condition)
df_summary <- df_bar %>%
  group_by(TreatmentGroup, Time_Point, Phylum) %>%
  summarise(RelAbundance = mean(RelAbundance), .groups = "drop")

# 7️⃣ Define color palette
phyla_all <- sort(unique(df_bar$Phylum))
n_colors <- max(3, min(12, length(phyla_all)))
palette_colors <- brewer.pal(n_colors, "Set3")
phylum_colors <- setNames(palette_colors[1:length(phyla_all)], phyla_all)

# 8️⃣ Plot the averaged bars (each bar = 100%)
p_bar <- ggplot(df_summary, aes(x = Time_Point, y = RelAbundance, fill = Phylum)) +
  geom_bar(stat = "identity", position = "stack", color = "black", linewidth = 0.2) +
  facet_wrap(~ TreatmentGroup) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Relative Taxonomic Composition at Phylum Level (T0, T2, T3, T6)",
    y = "Relative Abundance (%)",
    x = "Time Point"
  ) +
  scale_fill_manual(values = phylum_colors) +
  theme(
    plot.title = element_text(size = 20, face = "bold", margin = margin(b = 15)),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

print(p_bar)

# 9️⃣ Save plots
ggsave(file.path(root.dir, "phylum_barplot_relative_groupmean.pdf"), plot = p_bar, width = 10, height = 6)
ggsave(file.path(root.dir, "phylum_barplot_relative_groupmean.tiff"), plot = p_bar, width = 10, height = 6, dpi = 300)

```

Circular plots with F/B ratio

```{r donut phyla plots, fig.width=15, fig.height=10}
library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)  # for percent_format()
library(tibble)  # for rownames_to_column()
library(RColorBrewer)


# === 1. Filter samples for selected time points and prune taxa ===
included_timepoints <- c("T0", "T2", "T3", "T6")
ps_filtered <- subset_samples(ps, Time_Point %in% included_timepoints)
ps_filtered <- prune_taxa(taxa_sums(ps_filtered) > 0, ps_filtered)

# === 2. Agglomerate to Phylum level ===
ps_phylum <- tax_glom(ps_filtered, taxrank = "Phylum")

# === 3. Clean and fix taxonomy table ===
tax_mat <- as.matrix(tax_table(ps_phylum))
tax_mat[, "Phylum"] <- trimws(tax_mat[, "Phylum"])
tax_mat[, "Phylum"][tax_mat[, "Phylum"] == "none"] <- "Other"

# Replace tax table in phyloseq object
tax_table(ps_phylum) <- tax_table(tax_mat)

# === 4. Define phyla and color palette ===
phyla_all <- sort(unique(tax_mat[, "Phylum"]))
if (!"Other" %in% phyla_all) phyla_all <- c(phyla_all, "Other")

n_colors <- length(phyla_all)
n_colors <- max(3, min(12, n_colors))  # Brewer palettes support 3-12 colors

palette_colors <- brewer.pal(n_colors, "Set3")
phylum_colors <- setNames(palette_colors[1:length(phyla_all)], phyla_all)

# === 5. Prepare relative abundance data for plotting ===
otu_rel <- as(otu_table(ps_phylum), "matrix")
if (taxa_are_rows(ps_phylum)) otu_rel <- t(otu_rel)
otu_rel_df <- as.data.frame(otu_rel) %>% rownames_to_column("SampleID")

tax_df <- as.data.frame(tax_table(ps_phylum)) %>% rownames_to_column("FeatureID")

sample_df <- data.frame(sample_data(ps_phylum))
sample_df$SampleID <- rownames(sample_df)

# Join OTU relative abundance, taxonomy, and sample metadata
rel_abund_df <- otu_rel_df %>%
  pivot_longer(-SampleID, names_to = "FeatureID", values_to = "RelAbundance") %>%
  dplyr::left_join(tax_df %>% dplyr::select(FeatureID, Phylum), by = "FeatureID") %>%
  dplyr::left_join(sample_df %>% dplyr::select(SampleID, TreatmentGroup, Time_Point), by = "SampleID") %>%
  filter(!is.na(Phylum), !is.na(TreatmentGroup), !is.na(Time_Point))

# === 6. Summarize relative abundance per Phylum within each TreatmentGroup and Time_Point ===
library(dplyr)

# Define a threshold below which phyla are lumped into "Other"
threshold <- 0.005  # 1%

phylum_summary <- rel_abund_df %>%
  group_by(TreatmentGroup, Time_Point, Phylum) %>%
  summarise(RelAbundance = sum(RelAbundance), .groups = "drop") %>%
  group_by(TreatmentGroup, Time_Point) %>%
  # Calculate total per group for normalization
  mutate(total_group = sum(RelAbundance)) %>%
  # Calculate proportion
  mutate(prop = RelAbundance / total_group) %>%
  # Mark low abundance phyla as "Other"
  mutate(Phylum = ifelse(prop < threshold, "Other", Phylum)) %>%
  # Re-aggregate with "Other"
  group_by(TreatmentGroup, Time_Point, Phylum) %>%
  summarise(RelAbundance = sum(RelAbundance), .groups = "drop") %>%
  group_by(TreatmentGroup, Time_Point) %>%
  # Normalize again to make sure sums = 1
  mutate(RelAbundance = RelAbundance / sum(RelAbundance)) %>%
  arrange(TreatmentGroup, Time_Point, desc(RelAbundance)) %>%
  mutate(
    ymax = cumsum(RelAbundance),
    ymin = ymax - RelAbundance,
    label_pos = (ymax + ymin) / 2
  )


# === 7. Fix factor levels to match color names ===
phylum_summary$Phylum <- factor(phylum_summary$Phylum, levels = names(phylum_colors))

# === 8. Plot faceted donut chart with matching colors ===
p_donut <- ggplot(phylum_summary, aes(ymax = ymax, ymin = ymin, xmax = 4, xmin = 3, fill = Phylum)) +
  geom_rect(color = "black", size = 0.3) +   # black outline, adjust size as needed
  coord_polar(theta = "y") +
  facet_grid(TreatmentGroup ~ Time_Point, switch = "y") +  # switch strips to left for rows
  xlim(c(2, 4)) +
  theme_void() +
  theme(
    strip.placement = "outside",       # place facet labels outside the plot area
    strip.background = element_blank(), # remove strip background
    strip.text.y.left = element_text(angle = 0, hjust = 1, size = 11, face = "bold"), # style left strips
    legend.position = "right"
  ) +
  scale_fill_manual(values = phylum_colors) +
  labs(title = "Relative Abundance of Phyla by Treatment and Time Point")

print(p_donut)


# === 9. Save plots ===
ggsave(file.path(root.dir, "phylum_donut_plots.pdf"), plot = p_donut, width = 12, height = 8)
ggsave(file.path(root.dir, "phylum_donut_plots.tiff"), plot = p_donut, width = 12, height = 8, dpi = 300)



```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

# Step 1: Prepare taxonomy table with trimmed Phylum names
tax_df <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("FeatureID") %>%
  mutate(Phylum = trimws(as.character(Phylum)))

# Step 2: Filter Firmicutes and Bacteroidota features
firm_bact_ids <- tax_df %>%
  filter(Phylum %in% c("Firmicutes", "Bacteroidota")) %>%
  dplyr::select(FeatureID, Phylum)


# Step 3: Prepare OTU table in long format
otu_df <- as.data.frame(otu_table(ps))
if (taxa_are_rows(ps)) {
  otu_df <- t(otu_df)
}
otu_df <- as.data.frame(otu_df) %>%
  rownames_to_column("SampleID") %>%
  pivot_longer(-SampleID, names_to = "FeatureID", values_to = "Abundance")

# Step 4: Join taxonomy to OTU data, keep only Firmicutes and Bacteroidota
otu_tax_df <- inner_join(otu_df, firm_bact_ids, by = "FeatureID")

# Step 5: Summarize abundance per SampleID and Phylum, pivot wider
otu_summarized <- otu_tax_df %>%
  group_by(SampleID, Phylum) %>%
  summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
  pivot_wider(names_from = Phylum, values_from = TotalAbundance, values_fill = 0) %>%
  mutate(FB_Ratio = Firmicutes / Bacteroidota)

# Step 6: Add sample metadata
sample_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")
otu_summarized <- left_join(otu_summarized, sample_df, by = "SampleID")

# Step 7: Plot Firmicutes:Bacteroidota ratio over time by TreatmentGroup
# Make sure Time_Point is an ordered factor if you want ordered x-axis
otu_summarized$Time_Point <- factor(otu_summarized$Time_Point, levels = c("T0", "T2", "T3", "T4", "T6", "T8"))

p <- ggplot(otu_summarized, aes(x = Time_Point, y = FB_Ratio, color = TreatmentGroup)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "line", aes(group = TreatmentGroup), size = 1.2) +
  labs(
    title = "Firmicutes to Bacteroidota Ratio Over Time",
    x = "Time Point",
    y = "Firmicutes / Bacteroidota Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 25, margin = margin(b = 15)),  # bigger title & margin below
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    legend.title = element_text(size = 25),
    legend.key.height = unit(1.8, "lines"),   # increase spacing here# legend title size
    legend.text = element_text(size = 22)        # legend item text size
  )



print(p)

# Save plots
ggsave(filename = file.path(root.dir, "Firmicutes_Bacteroidota_Ratio.pdf"), plot = p, device = "pdf", width = 30, height = 20)
ggsave(filename = file.path(root.dir, "Firmicutes_Bacteroidota_Ratio.tiff"), plot = p, device = "tiff", dpi = 300, width = 10, height = 6)



```
Filter for specific groups
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)

# Step 1: Prepare taxonomy table with trimmed Phylum names
tax_df <- as.data.frame(tax_table(ps)) %>%
  rownames_to_column("FeatureID") %>%
  mutate(Phylum = trimws(as.character(Phylum)))

# Step 2: Filter Firmicutes and Bacteroidota features
firm_bact_ids <- tax_df %>%
  filter(Phylum %in% c("Firmicutes", "Bacteroidota")) %>%
  dplyr::select(FeatureID, Phylum)

# Step 3: Prepare OTU table in long format
otu_df <- as.data.frame(otu_table(ps))
if (taxa_are_rows(ps)) {
  otu_df <- t(otu_df)
}
otu_df <- as.data.frame(otu_df) %>%
  rownames_to_column("SampleID") %>%
  pivot_longer(-SampleID, names_to = "FeatureID", values_to = "Abundance")

# Step 4: Join taxonomy to OTU data, keep only Firmicutes and Bacteroidota
otu_tax_df <- inner_join(otu_df, firm_bact_ids, by = "FeatureID")

# Step 5: Summarize abundance per SampleID and Phylum, pivot wider
otu_summarized <- otu_tax_df %>%
  group_by(SampleID, Phylum) %>%
  summarise(TotalAbundance = sum(Abundance), .groups = "drop") %>%
  pivot_wider(names_from = Phylum, values_from = TotalAbundance, values_fill = 0) %>%
  mutate(FB_Ratio = Firmicutes / Bacteroidota)

# Step 6: Add sample metadata
sample_df <- data.frame(sample_data(ps)) %>%
  rownames_to_column("SampleID")
otu_summarized <- left_join(otu_summarized, sample_df, by = "SampleID")

# Step 7: Filter for specific TreatmentGroups and Time_Points
otu_summarized <- otu_summarized %>%
  filter(TreatmentGroup %in% c("Vehicle", "Olanzapine"),
         Time_Point %in% c("T0", "T2", "T3", "T6"))

# Step 8: Make sure Time_Point is a factor with desired order for plotting
otu_summarized$Time_Point <- factor(otu_summarized$Time_Point, levels = c("T0", "T2", "T3", "T6"))

# Step 9: Plot Firmicutes:Bacteroidota ratio over time by TreatmentGroup
p <- ggplot(otu_summarized, aes(x = Time_Point, y = FB_Ratio, color = TreatmentGroup)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
  stat_summary(fun = mean, geom = "line", aes(group = TreatmentGroup), size = 1.2) +
  labs(
    title = "Olanzapine vs Vehicle Firmicutes to Bacteroidota Ratio Over Time",
    x = "Time Point",
    y = "Firmicutes / Bacteroidota Ratio"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 25, margin = margin(b = 15)),  # bigger title & margin below
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right",
    legend.title = element_text(size = 25),
    legend.key.height = unit(1.8, "lines"),   # increase spacing here
    legend.text = element_text(size = 22)
  )

print(p)

# Save plots
ggsave(filename = file.path(root.dir, "Firmicutes_Bacteroidota_Ratio_OlanzapineVSVehicle.pdf"), plot = p, device = "pdf", width = 30, height = 20)
ggsave(filename = file.path(root.dir, "Firmicutes_Bacteroidota_Ratio_OlanzapineVSVehicle.tiff"), plot = p, device = "tiff", dpi = 300, width = 30, height = 20)

```


# Relative abundance bar charts at multiple taxonomic ranks
# Requires: ps (phyloseq object), root.dir defined
```{r}
library(phyloseq)
library(tidyverse)
library(ggplot2)

# Ranks to plot
rank_vec <- c("Phylum", "Class", "Order", "Family", "Genus")

# Output directory
ra_dir <- file.path(root.dir, "Relative_Abundance_Barplots")
if (!dir.exists(ra_dir)) dir.create(ra_dir, recursive = TRUE)

# Make sure these sample vars exist (otherwise remove from aes below)
sample_data(ps)$TreatmentGroup <- factor(sample_data(ps)$TreatmentGroup)
sample_data(ps)$Time_Point     <- factor(sample_data(ps)$Time_Point)

for (rk in rank_vec) {
  message("Processing rank: ", rk)

  # Skip if rank not present in taxonomy table
  if (!(rk %in% colnames(tax_table(ps)))) {
    message("  - Rank ", rk, " not found in tax_table(ps); skipping.")
    next
  }

  # Agglomerate to this taxonomic rank
  ps_rank <- tax_glom(ps, taxrank = rk)
  ps_rank <- prune_taxa(taxa_sums(ps_rank) > 0, ps_rank)

  # Convert to relative abundance
  ps_rel <- transform_sample_counts(ps_rank, function(x) x / sum(x))

  # Extract data to long format
  phy_df <- psmelt(ps_rel)  # phyloseq -> data.frame
  # phy_df columns include: Sample, Abundance, rk, TreatmentGroup, Time_Point, etc.

  # Clean taxon names: replace NA / empty / "none" / "unclassified" with "Unclassified"
  phy_df[[rk]] <- as.character(phy_df[[rk]])
  phy_df[[rk]][is.na(phy_df[[rk]]) |
               phy_df[[rk]] == "" |
               tolower(phy_df[[rk]]) %in% c("none", "unclassified", "unknown")] <- "Unclassified"
  
  # Optionally group low-abundance taxa into "Other"
  # Here we define low-abundance taxa by mean rel. abundance across all samples
  tax_means <- phy_df %>%
    group_by(.data[[rk]]) %>%
    summarise(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop")

  # Keep top N taxa, others -> "Other"
  topN <- 15
  top_taxa <- tax_means %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = topN) %>%
    pull(.data[[rk]])

  phy_df[[rk]] <- ifelse(phy_df[[rk]] %in% top_taxa, phy_df[[rk]], "Other")
  phy_df[[rk]] <- factor(phy_df[[rk]])

  # Order samples by TreatmentGroup and Time_Point for nicer plotting
  phy_df <- phy_df %>%
    arrange(TreatmentGroup, Time_Point, Sample)

  phy_df$Sample <- factor(phy_df$Sample, levels = unique(phy_df$Sample))

  # Build bar plot: each bar = sample, stacked by taxon at this rank
  p_ra <- ggplot(phy_df, aes(x = Sample, y = Abundance, fill = .data[[rk]])) +
    geom_bar(stat = "identity") +
    facet_grid(~ Time_Point, scales = "free_x", space = "free_x") +
    labs(
      title = paste("Relative Abundance at", rk, "Level"),
      x = "Sample (grouped by Time_Point)",
      y = "Relative Abundance",
      fill = rk
    ) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
      strip.background = element_rect(fill = "grey90"),
      panel.spacing = unit(0.2, "lines")
    )

  # Save plots
  out_pdf <- file.path(ra_dir, paste0("Relative_Abundance_", rk, ".pdf"))
  out_png <- file.path(ra_dir, paste0("Relative_Abundance_", rk, ".png"))

  ggsave(out_pdf, plot = p_ra, width = 12, height = 6)
  ggsave(out_png, plot = p_ra, width = 12, height = 6, dpi = 300)
}
message("✅ Relative abundance barplots saved in: ", ra_dir)
```

# Relative abundance bar charts at multiple taxonomic ranks
# Averaged across samples within TreatmentGroup x Time_Point
```{r}
library(phyloseq)
library(tidyverse)
library(ggplot2)

rank_vec <- c("Phylum", "Class", "Order", "Family", "Genus")

ra_dir <- file.path(root.dir, "Relative_Abundance_Barplots_Averaged")
if (!dir.exists(ra_dir)) dir.create(ra_dir, recursive = TRUE)

sample_data(ps)$TreatmentGroup <- factor(sample_data(ps)$TreatmentGroup)
sample_data(ps)$Time_Point     <- factor(sample_data(ps)$Time_Point)

for (rk in rank_vec) {
  message("Processing rank: ", rk)

  if (!(rk %in% colnames(tax_table(ps)))) {
    message("  - Rank ", rk, " not found in tax_table(ps); skipping.")
    next
  }

  # Agglomerate to rank
  ps_rank <- tax_glom(ps, taxrank = rk)
  ps_rank <- prune_taxa(taxa_sums(ps_rank) > 0, ps_rank)

  # Relative abundance per sample
  ps_rel <- transform_sample_counts(ps_rank, function(x) x / sum(x))

  phy_df <- psmelt(ps_rel)

  # Clean taxon names
  phy_df[[rk]] <- as.character(phy_df[[rk]])
  phy_df[[rk]][is.na(phy_df[[rk]]) |
               phy_df[[rk]] == "" |
               tolower(phy_df[[rk]]) %in% c("none", "unclassified", "unknown")] <- "Unclassified"

  # --- choose top taxa and relabel others as "Other" ---
  tax_means <- phy_df %>%
    group_by(.data[[rk]]) %>%
    summarise(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop")

  topN <- 15
  top_taxa <- tax_means %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = topN) %>%
    pull(.data[[rk]])

  phy_df[[rk]] <- ifelse(phy_df[[rk]] %in% top_taxa, phy_df[[rk]], "Other")

  # --- CRITICAL STEP: collapse within each sample after relabeling ---
  phy_df_collapsed <- phy_df %>%
    group_by(Sample, TreatmentGroup, Time_Point, .data[[rk]]) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")

  # Check that within each sample sums ~1
  # (optional sanity check)
  # phy_df_collapsed %>%
  #   group_by(Sample) %>%
  #   summarise(total = sum(Abundance)) %>%
  #   print(n = 10)

  # --- now average across samples within TreatmentGroup x Time_Point ---
  df_avg <- phy_df_collapsed %>%
    group_by(TreatmentGroup, Time_Point, .data[[rk]]) %>%
    summarise(
      mean_abund = mean(Abundance, na.rm = TRUE),
      n_samples  = n_distinct(Sample),
      .groups    = "drop"
    )

  df_avg$TreatmentGroup <- factor(df_avg$TreatmentGroup,
                                  levels = levels(sample_data(ps)$TreatmentGroup))
  df_avg$Time_Point <- factor(df_avg$Time_Point,
                              levels = levels(sample_data(ps)$Time_Point))

  # Stacked bars: one per TreatmentGroup, faceted by Time_Point
p_ra <- ggplot(df_avg, aes(x = TreatmentGroup, y = mean_abund, fill = .data[[rk]])) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Time_Point) +
  labs(
    title = paste("Relative Abundance at", rk, "Level (Averaged by Treatment x Time_Point)"),
    x = "Treatment Group",
    y = "Mean Relative Abundance",
    fill = rk
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  scale_fill_manual(values = NaturePastel_palette) +   # ← ADD THIS
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey90"),
    panel.spacing = unit(0.3, "lines")
  )


  out_pdf <- file.path(ra_dir, paste0("Relative_Abundance_Averaged_", rk, ".pdf"))
  out_png <- file.path(ra_dir, paste0("Relative_Abundance_Averaged_", rk, ".png"))

  ggsave(out_pdf, plot = p_ra, width = 10, height = 5)
  ggsave(out_png, plot = p_ra, width = 10, height = 5, dpi = 300)
}

message("✅ Averaged relative abundance barplots saved in: ", ra_dir)

```


## Top genera per treatment: time-course barplots + stats + p-value brackets
## Each plot: one genus, one treatment, bars over Time_Point, with Nature Pastel colors
```{r}
library(phyloseq)
library(tidyverse)
library(ggplot2)
library(rstatix)
library(ggpubr)

# Nature Pastel palette (20 colors)
nature_pastel <- c(
  "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e",
  "#e6ab02", "#a6761d", "#666666", "#1f78b4", "#b2df8a",
  "#fb9a99", "#fdbf6f", "#cab2d6", "#ffff99", "#b15928",
  "#8dd3c7", "#bebada", "#fb8072", "#80b1d3", "#fdb462"
)

# Basic factors
sample_data(ps)$TreatmentGroup <- factor(sample_data(ps)$TreatmentGroup)
sample_data(ps)$Time_Point     <- factor(sample_data(ps)$Time_Point)

# Output root
top_genus_root <- file.path(root.dir, "Top_Genus_Timecourse_WithStats")
if (!dir.exists(top_genus_root)) dir.create(top_genus_root, recursive = TRUE)

# Parameters
rank_name <- "Genus"
topN      <- 10   # how many top genera per treatment to analyze/plot

# Loop over treatments
treatments <- levels(sample_data(ps)$TreatmentGroup)

for (treat in treatments) {
  message("===== Treatment: ", treat, " =====")

  # Directory for this treatment
  treat_dir <- file.path(top_genus_root, paste0("Treatment_", treat))
  if (!dir.exists(treat_dir)) dir.create(treat_dir, recursive = TRUE)

  # Subset to this treatment
  ps_treat <- subset_samples(ps, TreatmentGroup == treat)
  ps_treat <- prune_taxa(taxa_sums(ps_treat) > 0, ps_treat)

  if (nsamples(ps_treat) < 2) {
    message("  - Not enough samples; skipping.")
    next
  }

  # Agglomerate to Genus
  ps_genus <- tax_glom(ps_treat, taxrank = rank_name)
  ps_genus <- prune_taxa(taxa_sums(ps_genus) > 0, ps_genus)

  # Relative abundance
  ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))

  # Long data
  df <- psmelt(ps_rel)   # columns: Sample, Abundance, Genus, Time_Point, TreatmentGroup, ...

  # Clean genus names
  df[[rank_name]] <- as.character(df[[rank_name]])
  df[[rank_name]][is.na(df[[rank_name]]) |
                  df[[rank_name]] == "" |
                  tolower(df[[rank_name]]) %in% c("none", "unclassified", "unknown")] <- "Unclassified"

  # Choose top genera in this treatment (by mean abundance across all samples/timepoints)
  top_genera <- df %>%
    group_by(.data[[rank_name]]) %>%
    summarise(mean_abund = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_abund)) %>%
    slice_head(n = topN) %>%
    pull(.data[[rank_name]])

  if (length(top_genera) == 0) {
    message("  - No genera to analyze; skipping.")
    next
  }

  # Containers for plots and stats
  plot_list <- list()
  stats_list <- list()

  # Loop over top genera
  for (i in seq_along(top_genera)) {
    genus_i <- top_genera[i]
    message("  * Genus: ", genus_i)

    df_g <- df %>%
      filter(.data[[rank_name]] == genus_i)

    # If genus never appears (shouldn't happen), skip
    if (nrow(df_g) == 0) next

    # Summary for bar heights (mean ± SEM) per time point
    summary_g <- df_g %>%
      group_by(Time_Point) %>%
      summarise(
        mean_abund = mean(Abundance, na.rm = TRUE),
        sd_abund   = sd(Abundance, na.rm = TRUE),
        n          = n(),
        se_abund   = sd_abund / sqrt(n),
        .groups    = "drop"
      )

    # ---- Statistics across time points ----
    # Kruskal–Wallis test (robust for non-normal rel. abundances)
    kw <- tryCatch(
      df_g %>% kruskal_test(Abundance ~ Time_Point),
      error = function(e) NULL
    )

    # Pairwise Wilcoxon test if KW is significant at 0.05
    pwc_sig <- NULL
    if (!is.null(kw) && nrow(kw) > 0 && !is.na(kw$p) && kw$p < 0.05) {
      pwc <- df_g %>%
        pairwise_wilcox_test(
          Abundance ~ Time_Point,
          p.adjust.method = "BH"
        )

      pwc_sig <- pwc %>%
        filter(p.adj < 0.05) %>%
        arrange(p.adj)

      if (nrow(pwc_sig) > 0) {
        # Limit number of brackets to avoid clutter (e.g., top 4)
        pwc_sig <- pwc_sig %>%
          slice_head(n = 4)

        # Compute y positions for brackets
        y_max <- max(summary_g$mean_abund, na.rm = TRUE)
        # small offsets above the max bar
        pwc_sig <- pwc_sig %>%
          mutate(
            y.position = y_max + seq(0.02, 0.02 * n(), by = 0.02),
            label = paste0("p = ", signif(p.adj, 2)),
            group = row_number()
          )
      }
      # Save pairwise stats
      if (nrow(pwc) > 0) {
        stats_list[[length(stats_list) + 1]] <- pwc %>%
          mutate(
            Treatment   = treat,
            Genus       = genus_i,
            global_kw_p = kw$p
          )
      }
    } else {
      # Even if KW not sig, record its p
      stats_list[[length(stats_list) + 1]] <- tibble(
        Treatment   = treat,
        Genus       = genus_i,
        group1      = NA_character_,
        group2      = NA_character_,
        p           = NA_real_,
        p.adj       = NA_real_,
        p.adj.signif = NA_character_,
        method      = "Kruskal-Wallis",
        global_kw_p = if (!is.null(kw)) kw$p else NA_real_
      )
    }

    # ---- Choose a consistent color for this genus from Nature Pastel ----
    color_idx <- (i - 1) %% length(nature_pastel) + 1
    genus_color <- nature_pastel[color_idx]
    fill_vals <- setNames(genus_color, genus_i)

    # ---- Plot: mean ± SE over time, one bar per timepoint ----
    p <- ggplot(summary_g,
                aes(x = Time_Point, y = mean_abund, fill = genus_i)) +
      geom_col(color = "black", width = 0.7) +
      geom_errorbar(aes(ymin = pmax(mean_abund - se_abund, 0),
                        ymax = mean_abund + se_abund),
                    width = 0.2) +
      scale_fill_manual(values = fill_vals, guide = "none") +
      scale_y_continuous(
        labels = scales::percent_format(accuracy = 1),
        limits = c(0, min(1, max(summary_g$mean_abund + summary_g$se_abund, na.rm = TRUE) + 0.1))
      ) +
      labs(
        title = paste0("Treatment: ", treat,
                       " | Genus: ", genus_i),
        x = "Time Point",
        y = "Mean Relative Abundance"
      ) +
      theme_bw() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)
      )

    # Add p-value brackets, if any significant pairs
    if (!is.null(pwc_sig) && nrow(pwc_sig) > 0) {
      p <- p +
        ggpubr::stat_pvalue_manual(
          pwc_sig,
          label = "label",
          y.position = "y.position",
          xmin = "group1",
          xmax = "group2",
          tip.length = 0.01,
          bracket.size = 0.3,
          size = 3
        )
    }

    plot_list[[length(plot_list) + 1]] <- p
  }

  # ---- Save multi-page PDF of all genus plots for this treatment ----
  if (length(plot_list) > 0) {
    pdf(file.path(treat_dir, "Top_Genera_Timecourse_with_stats.pdf"),
        width = 7, height = 5)
    for (p in plot_list) print(p)
    dev.off()
  }

  # ---- Save stats CSV for this treatment ----
  if (length(stats_list) > 0) {
    stats_treat <- bind_rows(stats_list)
    write.csv(
      stats_treat,
      file = file.path(treat_dir, "Top_Genera_Timecourse_stats.csv"),
      row.names = FALSE
    )
  }
}

message("✅ Finished: per-treatment genus time-course plots + stats saved under:\n",
        top_genus_root, "\n")
```

###Below for Functional Predictions from 16S

```{r}
library(phyloseq)
library(Biostrings)

out_dir <- "PICRUSt2_from_phyloseq"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Make sure taxa are rows
if (!taxa_are_rows(ps)) ps <- t(ps)

# 1A) FASTA (headers = ASV IDs). Use safe, short IDs (PICRUSt2-friendly).
asv_ids <- paste0("ASV_", seq_len(ntaxa(ps)))
seqs <- taxa_names(ps)

fasta_path <- file.path(out_dir, "asv_rep_seqs.fasta")
con <- file(fasta_path, "w")
for (i in seq_along(seqs)) {
  writeLines(paste0(">", asv_ids[i]), con)
  writeLines(seqs[i], con)
}
close(con)

# 1B) Count table (samples x ASVs) with matching ASV IDs as column names
asv_mat <- as(otu_table(ps), "matrix")  # taxa x samples
rownames(asv_mat) <- asv_ids            # replace long DNA names with ASV_#
asv_mat <- t(asv_mat)                   # samples x taxa

asv_tsv <- file.path(out_dir, "asv_table.tsv")
write.table(asv_mat, asv_tsv, sep = "\t", quote = FALSE, col.names = NA)

# 1C) Save an ID→sequence key (handy for interpreting later)
key_path <- file.path(out_dir, "ASV_id_to_sequence.tsv")
write.table(data.frame(ASV_ID = asv_ids, Sequence = seqs),
            key_path, sep = "\t", quote = FALSE, row.names = FALSE)

cat("Wrote:\n", fasta_path, "\n", asv_tsv, "\n", key_path, "\n")

```

16s and metabolomics integration 

```{r}
# ============================================================
# STEP 1 — Define the single pre-challenge time point + subset
# Requires: ps already exists
# Output: ps_pre, meta_pre
# ============================================================

library(phyloseq)
library(dplyr)
library(tibble)

# --- set your pre-challenge time point label here ---
pre_tp <- "T6"   # change if needed (e.g., "T3")

# Subset to the one time point
ps_pre <- subset_samples(ps, Time_Point == pre_tp)
ps_pre <- prune_taxa(taxa_sums(ps_pre) > 0, ps_pre)

# Pull metadata (keep SampleID explicit)
meta_pre <- data.frame(sample_data(ps_pre), stringsAsFactors = FALSE) %>%
  rownames_to_column("SampleID") %>%
  mutate(
    SampleID = as.character(SampleID),
    TreatmentGroup = factor(TreatmentGroup),
    Time_Point = as.character(Time_Point)
  )

# Quick checks
cat("Time point kept:", pre_tp, "\n")
cat("N samples at pre-challenge:", nsamples(ps_pre), "\n")
print(table(meta_pre$TreatmentGroup))
stopifnot(all(meta_pre$SampleID %in% sample_names(ps_pre)))
```
